# The Two-Body Problem, Solved by Hand {#sec-two-body}

```{r}
#| include: false
source("_common.R")
```

The package documentation hands you a handful of formulas. The circular-orbit speed is $\sqrt{GM/r}$; escape speed is $\sqrt{2}$ times that; an orbit's period is $2\pi\sqrt{a^3/\mu}$; the distance from the parent at true anomaly $\nu$ is $a(1-e^2)/(1+e\cos\nu)$. Every one of them is true, and every one of them follows from Newton's law in a few lines of vector algebra. This chapter does the algebra. It is the one place in the book where the derivations come first and the code comes second, and it will repay the effort: these are the formulas you will use to set up every system in Part III.

The strategy is the physicist's standard one. First find the quantities that *do not change* as the bodies move, the conserved quantities. Then use them to pin down the shape of the orbit without ever integrating anything. Along the way, Kepler's three laws drop out as consequences.

## Two bodies become one

Take two point masses, $m_1$ at $\mathbf{r}_1$ and $m_2$ at $\mathbf{r}_2$, with nothing else in the universe. From @eq-nbody,

$$
\ddot{\mathbf{r}}_1 = G m_2 \frac{\mathbf{r}_2 - \mathbf{r}_1}{r^3}, \qquad
\ddot{\mathbf{r}}_2 = G m_1 \frac{\mathbf{r}_1 - \mathbf{r}_2}{r^3},
$$

where $r = |\mathbf{r}_2 - \mathbf{r}_1|$. Subtract the first from the second. Writing $\mathbf{r} = \mathbf{r}_2 - \mathbf{r}_1$ for the *relative* position of body 2 as seen from body 1,

$$
\ddot{\mathbf{r}} = -\frac{\mu}{r^3}\,\mathbf{r}, \qquad \mu = G(m_1 + m_2).
$$ {#eq-relative}

Two coupled equations in six coordinates have become one equation in three. It says the relative position vector accelerates toward the origin as if there were a single fixed mass $m_1 + m_2$ sitting there. The combination $\mu = G(m_1+m_2)$ is called the gravitational parameter, and it is the only combination of $G$ and the masses that the relative motion knows about.

What happened to the other three coordinates? Add $m_1$ times the first equation to $m_2$ times the second: the right-hand sides cancel, so the center of mass $\mathbf{R} = (m_1\mathbf{r}_1 + m_2\mathbf{r}_2)/(m_1+m_2)$ has $\ddot{\mathbf{R}} = 0$. It moves in a straight line at constant velocity, which is the drift we met in Chapter 2. Given $\mathbf{R}$ and $\mathbf{r}$, the individual positions are

$$
\mathbf{r}_1 = \mathbf{R} - \frac{m_2}{m_1+m_2}\,\mathbf{r}, \qquad
\mathbf{r}_2 = \mathbf{R} + \frac{m_1}{m_1+m_2}\,\mathbf{r}.
$$ {#eq-individual}

Both bodies trace the same shape of orbit around the center of mass, scaled by the mass ratio. In Chapter 2's fixed Earth-Moon system, Earth's little circle had a radius $m_{\mathrm{Moon}}/(m_{\mathrm{Earth}}+m_{\mathrm{Moon}}) \approx 1/82$ of the Moon's.

When one body is much heavier, $m_1 \gg m_2$, two things happen at once: $\mu \approx G m_1$, and $\mathbf{r}_1 \approx \mathbf{R}$, so the heavy body sits essentially at the center of mass and the light body orbits it. That is the "fixed central body" picture of the two-body vignette, and the approximation is excellent for the Sun and any planet (the Sun outweighs Jupiter a thousand to one) and merely good for Earth and the Moon, where the correction is 1.2%. From here on the derivation is exact for the relative motion; whether you then set $\mu = GM$ or $\mu = G(M+m)$ is a choice about how much precision you need.

## Angular momentum is conserved, and the orbit is flat

Define the *specific angular momentum* (angular momentum per unit mass) of the relative motion,

$$
\mathbf{h} = \mathbf{r} \times \dot{\mathbf{r}}.
$$

Differentiate it using the product rule:

$$
\dot{\mathbf{h}} = \dot{\mathbf{r}} \times \dot{\mathbf{r}} + \mathbf{r} \times \ddot{\mathbf{r}}
= \mathbf{0} + \mathbf{r} \times \left(-\frac{\mu}{r^3}\mathbf{r}\right) = \mathbf{0}.
$$

The first term is a vector crossed with itself and the second is $\mathbf{r}$ crossed with a multiple of $\mathbf{r}$; both vanish. So $\mathbf{h}$ is a constant of the motion. This works for *any* force directed along $\mathbf{r}$, not just inverse-square: it is a consequence of the force being central.

Two consequences follow immediately. Since $\mathbf{h}$ is perpendicular to both $\mathbf{r}$ and $\dot{\mathbf{r}}$ at every instant, and $\mathbf{h}$ never changes, the position and velocity are confined forever to the plane perpendicular to $\mathbf{h}$. **Every two-body orbit is planar.** And the magnitude $h = |\mathbf{r}\times\dot{\mathbf{r}}|$ is constant. In a small time $dt$ the position vector sweeps out a thin triangle with sides $\mathbf{r}$ and $\dot{\mathbf{r}}\,dt$, whose area is half the magnitude of their cross product:

$$
\frac{dA}{dt} = \frac{1}{2}\,|\mathbf{r} \times \dot{\mathbf{r}}| = \frac{h}{2}.
$$ {#eq-areal}

The line from the central body to the orbiting one sweeps out equal areas in equal times. That is **Kepler's second law**, and it took one line.

## Energy is conserved

Now dot @eq-relative with the velocity:

$$
\dot{\mathbf{r}}\cdot\ddot{\mathbf{r}} = -\frac{\mu}{r^3}\,\mathbf{r}\cdot\dot{\mathbf{r}}.
$$

The left side is $\frac{d}{dt}\left(\tfrac12 \dot{\mathbf{r}}\cdot\dot{\mathbf{r}}\right) = \frac{d}{dt}\left(\tfrac12 v^2\right)$. For the right side, note that $r^2 = \mathbf{r}\cdot\mathbf{r}$, so $2r\dot{r} = 2\,\mathbf{r}\cdot\dot{\mathbf{r}}$ and $\mathbf{r}\cdot\dot{\mathbf{r}} = r\dot{r}$. Then $-\mu\,\mathbf{r}\cdot\dot{\mathbf{r}}/r^3 = -\mu\dot{r}/r^2 = \frac{d}{dt}(\mu/r)$. Moving everything to one side,

$$
\frac{d}{dt}\left(\frac{v^2}{2} - \frac{\mu}{r}\right) = 0, \qquad
\varepsilon \equiv \frac{v^2}{2} - \frac{\mu}{r} = \text{const}.
$$ {#eq-energy}

$\varepsilon$ is the *specific orbital energy*: kinetic energy per unit mass plus gravitational potential energy per unit mass. If $\varepsilon < 0$ the body cannot reach $r = \infty$ (it would need negative kinetic energy there), so it is bound. If $\varepsilon \ge 0$ it can, and it escapes.

Chapter 12 turns both of these conservation laws into diagnostics: a simulation that lets $\varepsilon$ or $\mathbf{h}$ drift is a simulation with a problem.

## The orbit equation

Here is the step that makes the whole chapter work. Conservation of $\mathbf{h}$ and $\varepsilon$ constrains the motion but does not by itself give the shape of the path. For the inverse-square force, and essentially only for it, there is a *third* conserved vector, and it hands you the orbit directly.

Start from the time derivative of $\dot{\mathbf{r}}\times\mathbf{h}$. Since $\mathbf{h}$ is constant,

$$
\frac{d}{dt}\left(\dot{\mathbf{r}}\times\mathbf{h}\right) = \ddot{\mathbf{r}}\times\mathbf{h}
= -\frac{\mu}{r^3}\,\mathbf{r}\times(\mathbf{r}\times\dot{\mathbf{r}}).
$$

Expand the triple product with the identity $\mathbf{a}\times(\mathbf{b}\times\mathbf{c}) = \mathbf{b}(\mathbf{a}\cdot\mathbf{c}) - \mathbf{c}(\mathbf{a}\cdot\mathbf{b})$:

$$
\mathbf{r}\times(\mathbf{r}\times\dot{\mathbf{r}}) = \mathbf{r}\,(\mathbf{r}\cdot\dot{\mathbf{r}}) - \dot{\mathbf{r}}\,r^2
= \mathbf{r}\,r\dot{r} - \dot{\mathbf{r}}\,r^2,
$$

so

$$
\frac{d}{dt}\left(\dot{\mathbf{r}}\times\mathbf{h}\right)
= \mu\left(\frac{\dot{\mathbf{r}}}{r} - \frac{\dot{r}\,\mathbf{r}}{r^2}\right)
= \mu\,\frac{d}{dt}\left(\frac{\mathbf{r}}{r}\right).
$$

The last equality is just the quotient rule applied to $\mathbf{r}/r$. Both sides are total time derivatives, so their difference is a constant vector. Divide by $\mu$ and call the constant $\mathbf{e}$:

$$
\mathbf{e} = \frac{\dot{\mathbf{r}}\times\mathbf{h}}{\mu} - \frac{\mathbf{r}}{r}.
$$ {#eq-ecc-vector}

This is the *eccentricity vector* (also called the Laplace-Runge-Lenz vector, in various normalizations). It lies in the orbital plane, because both terms do. To see what it means, take its dot product with $\mathbf{r}$. The first term gives $\mathbf{r}\cdot(\dot{\mathbf{r}}\times\mathbf{h})/\mu = \mathbf{h}\cdot(\mathbf{r}\times\dot{\mathbf{r}})/\mu = h^2/\mu$, using the cyclic property of the scalar triple product; the second gives $-r$. If $\nu$ is the angle between $\mathbf{e}$ and $\mathbf{r}$, the dot product is also $e\,r\cos\nu$. So

$$
e\,r\cos\nu = \frac{h^2}{\mu} - r
\quad\Longrightarrow\quad
r = \frac{h^2/\mu}{1 + e\cos\nu}.
$$ {#eq-orbit}

This is the **orbit equation**: the distance from the central body as a function of the angle $\nu$ measured from the direction of $\mathbf{e}$. The quantity $p = h^2/\mu$ is called the semi-latus rectum, and $e = |\mathbf{e}|$ is the eccentricity. The distance $r$ is smallest when $\cos\nu = 1$, so $\mathbf{e}$ points toward the point of closest approach, the *periapsis*, and $\nu$, the *true anomaly*, is the angle from periapsis. @eq-orbit is a conic section with one focus at the origin:

| $e$ | $\varepsilon$ | shape | $r_{\max}$ |
|:--|:--|:--|:--|
| $0$ | $-\mu^2/2h^2$ | circle | $= r_{\min}$ |
| $0 < e < 1$ | $< 0$ | ellipse | finite |
| $1$ | $0$ | parabola | $\infty$ |
| $> 1$ | $> 0$ | hyperbola | $\infty$ |

: The four kinds of two-body orbit. {#tbl-conics}

Bound orbits are ellipses with the central body at a focus. That is **Kepler's first law**, and it has now been derived from the inverse-square law, which is the thing Halley asked Newton about in 1684.

The relation between $e$ and $\varepsilon$ in the table comes from squaring @eq-ecc-vector. Since $\dot{\mathbf{r}} \perp \mathbf{h}$, $|\dot{\mathbf{r}}\times\mathbf{h}| = vh$, and the cross term is $-2\,\mathbf{r}\cdot(\dot{\mathbf{r}}\times\mathbf{h})/(\mu r) = -2h^2/(\mu r)$, so

$$
e^2 = \frac{v^2 h^2}{\mu^2} - \frac{2h^2}{\mu r} + 1 = 1 + \frac{2\varepsilon h^2}{\mu^2}.
$$ {#eq-e-energy}

Negative energy means $e<1$: bound orbits are ellipses, as the table says.

::: {.callout-note title="Where it comes from"}
Kepler extracted his first two laws from Tycho Brahe's observations of Mars, published in *Astronomia Nova* in 1609 [@kepler1609], and the third from the periods and distances of all the planets, published in *Harmonices Mundi* in 1619 [@kepler1619]. He had no theory of why; he had tried circles and epicycles for years and the data would not fit. Newton's *Principia* [@newton1687] showed that all three laws follow from an inverse-square force (Book I, Propositions 11 to 15), using geometry rather than vectors. The vector form here is the standard modern one; Goldstein [@goldstein2002] and Bate, Mueller, and White [@bate1971] both present it, and the eccentricity vector is the starting point of every algorithm that turns a position and velocity into orbital elements, including the one in Chapter 7.
:::

## The geometry of an ellipse

An ellipse has a longest diameter, the major axis, of length $2a$; $a$ is the semi-major axis. From @eq-orbit the nearest and farthest points are at $\nu = 0$ and $\nu = \pi$:

$$
r_p = \frac{p}{1+e}, \qquad r_a = \frac{p}{1-e}, \qquad a = \frac{r_p + r_a}{2} = \frac{p}{1-e^2}.
$$

So $p = a(1-e^2)$, and the orbit equation takes the form the package uses:

$$
r = \frac{a(1-e^2)}{1+e\cos\nu}, \qquad r_p = a(1-e), \qquad r_a = a(1+e).
$$ {#eq-orbit-a}

The semi-minor axis is $b = a\sqrt{1-e^2}$. Two more facts fall out of these. First, given any two-body orbit, you can read off its eccentricity from its extreme distances alone:

$$
e = \frac{r_a - r_p}{r_a + r_p}.
$$ {#eq-e-from-extremes}

Second, substituting $h^2 = \mu p = \mu a(1-e^2)$ into @eq-e-energy and simplifying (try it; everything but one factor cancels) gives the energy in terms of the size of the orbit alone:

$$
\varepsilon = -\frac{\mu}{2a}.
$$ {#eq-energy-a}

The energy of an orbit depends only on its semi-major axis, not on its eccentricity. A circle of radius $a$ and a needle-thin ellipse of semi-major axis $a$ have the same energy.

## Vis-viva, circular speed, and escape speed

Put @eq-energy-a into the definition of $\varepsilon$ and solve for the speed:

$$
v^2 = \mu\left(\frac{2}{r} - \frac{1}{a}\right).
$$ {#eq-vis-viva}

This is the *vis-viva* equation (Latin for "living force," an old name for kinetic energy). It gives the speed at any distance $r$ along an orbit of semi-major axis $a$, and it is the formula you want whenever you know where a body is and need to know how fast it should be going.

Two special cases are the ones the documentation quotes. On a circular orbit, $r = a$ always, so

$$
v_{\mathrm{circ}} = \sqrt{\frac{\mu}{r}}.
$$ {#eq-v-circ}

On a parabolic orbit, $a \to \infty$ and the $1/a$ term vanishes, so the speed that just barely escapes from distance $r$ is

$$
v_{\mathrm{esc}} = \sqrt{\frac{2\mu}{r}} = \sqrt{2}\,v_{\mathrm{circ}}.
$$ {#eq-v-esc}

Now the rules of thumb in the two-body vignette can be made exact. Suppose you place a body at distance $r$ from the central mass and launch it perpendicular to the radius at speed $v = k\,v_{\mathrm{circ}}$. Then $h = rv$, so $p = h^2/\mu = r^2v^2/\mu = k^2 r$. The launch point has $\dot{r} = 0$, so it is either periapsis or apoapsis, and @eq-orbit gives $r = p/(1\pm e)$, or

$$
e = \left|k^2 - 1\right|.
$$ {#eq-e-from-k}

If $k > 1$ the launch point is periapsis and the orbit swings outward; if $k < 1$ it is apoapsis and the orbit dips inward; $k = 1$ is a circle; $k = \sqrt{2}$ gives $e = 1$, a parabola. A planet launched at 1.2 times circular speed has $e = 0.44$; at 1.4 times, $e = 0.96$, a sliver away from escape. We will check both numbers in a moment.

## Kepler's third law

The area of an ellipse is $\pi a b$. By @eq-areal the orbit sweeps area at the constant rate $h/2$, so one full orbit takes $T = 2\pi ab/h$. Substitute $b = a\sqrt{1-e^2}$ and $h = \sqrt{\mu a (1-e^2)}$:

$$
T = \frac{2\pi a^2\sqrt{1-e^2}}{\sqrt{\mu a(1-e^2)}} = 2\pi\sqrt{\frac{a^3}{\mu}}.
$$ {#eq-kepler3}

The eccentricity has cancelled: **Kepler's third law**. The period depends on the semi-major axis and the total mass and on nothing else. Kepler's own statement, that $T^2/a^3$ is the same number for every planet, holds only because every planet's mass is negligible next to the Sun's, so $\mu \approx GM_\odot$ for all of them. The correction for Jupiter is one part in a thousand in $\mu$, half that in $T$.

This is also the law that weighs things. Measure a moon's period and orbital radius and @eq-kepler3 gives you the mass of its planet; this is how the masses in @sec-constants were obtained, and it is how we know the mass of the Sun, of Jupiter, and of the black hole at the center of the Milky Way.

## Checking everything against the simulation

Time to see whether the package agrees with the algebra. The derivations above are exact for two bodies; the simulation is approximate; if they match, both the physics and the numerics are in good shape. To keep the comparison honest I will compute every prediction with the full $\mu = G(M+m)$.

### The speed sweep

Set up the vignette's star and planet, and launch the planet at several multiples of circular speed:

```{r}
M  <- 1e30
m  <- 1e24
r0 <- 1e11
mu <- gravitational_constant * (M + m)
v_circ <- sqrt(mu / r0)

make_sim <- function(k, duration = seconds_per_year * 2) {
  create_system() |>
    add_body("Star",   mass = M) |>
    add_body("Planet", mass = m, x = r0, vy = k * v_circ) |>
    simulate_system(time_step = seconds_per_hour, duration = duration) |>
    mutate(k = k)
}

sweep <- bind_rows(lapply(c(0.7, 1.0, 1.2, 1.4), make_sim))
```

```{r}
#| label: fig-speed-sweep
#| fig-cap: "The same launch point, four launch speeds. The circle is $k = 1$; slower launches dip inward from apoapsis, faster ones swing outward from periapsis. The $k = 1.4$ orbit has $e = 0.96$ and a period of nearly a century, so two years show only its beginning."
#| fig-width: 5
#| fig-height: 4.6
#| out-width: "80%"
sweep |>
  filter(id == "Planet") |>
  ggplot(aes(x, y, color = factor(k))) +
  geom_path() +
  annotate("point", x = 0, y = 0, size = 3) +
  coord_equal() +
  labs(x = NULL, y = NULL, color = expression(v / v[circ]))
```

### Measuring eccentricity from a single state

@eq-ecc-vector is more than a derivation device. It gives the eccentricity from the position and velocity at *any one instant*, no full orbit required. Two helpers make that easy. The first extracts the relative motion of one body about another, the quantity all of this chapter's formulas are about:

```{r}
relative_state <- function(sim, body, center, G = gravitational_constant) {
  ctr <- sim |>
    filter(id == center) |>
    select(time, m_c = mass, x_c = x, y_c = y, z_c = z,
           vx_c = vx, vy_c = vy, vz_c = vz)
  sim |>
    filter(id == body) |>
    inner_join(ctr, by = "time") |>
    mutate(mu = G * (mass + m_c),
           rx = x - x_c,   ry = y - y_c,   rz = z - z_c,
           ux = vx - vx_c, uy = vy - vy_c, uz = vz - vz_c) |>
    select(time, mu, rx, ry, rz, ux, uy, uz)
}
```

The second computes $\mathbf{h} = \mathbf{r}\times\mathbf{u}$ and then $\mathbf{e} = (\mathbf{u}\times\mathbf{h})/\mu - \mathbf{r}/r$ component by component:

```{r}
eccentricity <- function(sim, body, center, G = gravitational_constant) {
  relative_state(sim, body, center, G) |>
    mutate(r  = sqrt(rx^2 + ry^2 + rz^2),
           hx = ry * uz - rz * uy,                   # h = r x u
           hy = rz * ux - rx * uz,
           hz = rx * uy - ry * ux,
           ex = (uy * hz - uz * hy) / mu - rx / r,  # e = (u x h)/mu - r/r
           ey = (uz * hx - ux * hz) / mu - ry / r,
           ez = (ux * hy - uy * hx) / mu - rz / r,
           e  = sqrt(ex^2 + ey^2 + ez^2)) |>
    select(time, e, ex, ey, ez)
}
```

Now compare the prediction $e = |k^2 - 1|$ with the eccentricity measured from the simulated state at the *end* of each run, two years of integration away from the initial conditions:

```{r}
sweep |>
  group_by(k) |>
  group_modify(~ eccentricity(.x, "Planet", "Star") |>
                 filter(time == max(time))) |>
  ungroup() |>
  mutate(predicted = abs(k^2 - 1)) |>
  select(k, predicted, measured = e)
```

The agreement is to many decimal places. And since $\mathbf{e}$ is supposed to be a constant of the motion, we can ask how constant the simulation kept it over the whole two years for the $k = 1.2$ orbit:

```{r}
e_track <- eccentricity(filter(sweep, k == 1.2), "Planet", "Star")
range(e_track$e)
```

The spread between the smallest and largest value is the integrator's error. It is small. Chapter 5 explains why it is small and what it depends on, and shows what the same test looks like with a worse integrator.

The package does this computation for you. `get_orbital_elements()` applies exactly the eccentricity-vector and vis-viva formulas of this chapter to every time step and returns the full set of orbital elements, with the semi-major axis and the Kepler's-third-law period alongside the eccentricity:

```{r}
get_orbital_elements(filter(sweep, k == 1.2), "Planet", "Star",
                     mu = gravitational_constant * (M + m)) |>
  summarise(e = mean(e), a = mean(a), period_days = mean(period) / seconds_per_day)
```

(The `mu` argument is there because, by default, the function uses the parent's mass alone, which is the convention of `add_body_keplerian()` and is exact in the test-particle limit; passing $G(M+m)$ gives the exact two-body elements, which is what the derivations here are about.) I wrote `eccentricity()` out by hand above so that you could see the formula do its work; from here on the book uses the package function, and Chapter 7 derives the rest of what it computes.

### Measuring the period

Kepler's third law needs a period. `get_orbital_elements()` reports one, but it is computed *from* the third law, $T = 2\pi\sqrt{a^3/\mu}$, so using it to check the third law would be circular. An independent measurement means finding when the body has actually gone all the way round. The cleanest way is to track the polar angle of the relative position, unwrap the jump from $+\pi$ to $-\pi$, and interpolate to the moment the total angle reaches $2\pi$:

```{r}
measure_period <- function(sim, body, center = NULL, G = gravitational_constant) {
  if (is.null(center)) {                              # measure about the origin
    rel <- sim |> filter(id == body) |> transmute(time, rx = x, ry = y)
  } else {
    rel <- relative_state(sim, body, center, G)
  }
  theta  <- atan2(rel$ry, rel$rx)
  dtheta <- ((diff(theta) + pi) %% (2 * pi)) - pi   # unwrap the jump at +/- pi
  swept  <- abs(cumsum(c(0, dtheta)))                # total angle travelled
  i <- which(swept >= 2 * pi)[1]
  if (is.na(i)) return(NA_real_)                     # the run was too short
  f <- (2 * pi - swept[i - 1]) / (swept[i] - swept[i - 1])
  rel$time[i - 1] + f * (rel$time[i] - rel$time[i - 1])
}
```

For the circular case, @eq-kepler3 with $a = r_0$:

```{r}
T_pred <- 2 * pi * sqrt(r0^3 / mu)
T_meas <- measure_period(filter(sweep, k == 1), "Planet", "Star")

c(predicted_days = T_pred / seconds_per_day,
  measured_days  = T_meas / seconds_per_day)
```

For the $k = 1.2$ ellipse the semi-major axis is $a = r_p/(1-e)$ with $r_p = r_0$ and $e = 0.44$, which makes the orbit larger and the period longer:

```{r}
a_12   <- r0 / (1 - 0.44)
T_pred <- 2 * pi * sqrt(a_12^3 / mu)
T_meas <- measure_period(filter(sweep, k == 1.2), "Planet", "Star")

c(predicted_days = T_pred / seconds_per_day,
  measured_days  = T_meas / seconds_per_day)
```

### The Moon, revisited

Chapter 2's four-line Moon started at the semi-major-axis distance of 384,400 km with the mean orbital speed of 1,022 m/s. We can now say exactly what orbit that is. The circular speed at that distance, with the full $\mu = G(M_{\mathrm{Earth}} + m_{\mathrm{Moon}})$, is a little *higher* than 1,022 m/s, so the launch point is apoapsis and the orbit dips inward:

```{r}
mu_em  <- gravitational_constant * (mass_earth + mass_moon)
k_moon <- speed_moon / sqrt(mu_em / distance_earth_moon)
e_moon <- abs(k_moon^2 - 1)
a_moon <- distance_earth_moon / (1 + e_moon)      # launch point is apoapsis
T_moon <- 2 * pi * sqrt(a_moon^3 / mu_em)

c(k = k_moon, e = e_moon, a_km = a_moon / 1e3, T_days = T_moon / seconds_per_day)
```

Against the simulation:

```{r}
moon <- create_system() |>
  add_body("Earth", mass = mass_earth) |>
  add_body("Moon",  mass = mass_moon, x = distance_earth_moon, vy = speed_moon) |>
  simulate_system(time_step = seconds_per_hour, duration = seconds_per_day * 28)

measure_period(moon, "Moon", "Earth") / seconds_per_day
```

The simulation and the formula agree. But the real Moon's sidereal period is 27.32 days, and the four-line Moon's is about `r round(T_moon / seconds_per_day, 1)`. The simulation is not wrong; the *setup* describes a slightly different orbit from the real one. "Semi-major axis" and "mean speed" are two properties of the real elliptical orbit, but using one as a starting distance and the other as a starting speed does not reproduce that ellipse; it produces a nearby one with a smaller $a$ and hence a shorter period. The right way to put a body on its real orbit is to specify the orbit itself, by its orbital elements, and let the package compute a consistent position and velocity. That is what `add_planet()` does and what Chapter 7 derives.

::: {.callout-tip title="Key results"}
For the relative motion of two bodies with $\mu = G(m_1 + m_2)$:

- $\mathbf{h} = \mathbf{r}\times\dot{\mathbf{r}}$ and $\varepsilon = v^2/2 - \mu/r$ are constant; the orbit is planar and sweeps equal areas in equal times.
- Orbit equation: $r = a(1-e^2)/(1+e\cos\nu)$, with the central body at a focus. $e<1$ bound, $e \ge 1$ unbound.
- $\varepsilon = -\mu/2a$, independent of $e$.
- Vis-viva: $v^2 = \mu(2/r - 1/a)$. Circular speed $\sqrt{\mu/r}$; escape speed $\sqrt{2\mu/r}$.
- Launching perpendicular to the radius at $k$ times circular speed gives $e = |k^2-1|$.
- Period: $T = 2\pi\sqrt{a^3/\mu}$.
- $e = (r_a - r_p)/(r_a + r_p)$, and $\mathbf{e} = (\dot{\mathbf{r}}\times\mathbf{h})/\mu - \mathbf{r}/r$ from any single state.
:::
