# Integrators: Euler, Euler-Cromer, and Velocity Verlet {#sec-integrators}

```{r}
#| include: false
source("_common.R")
```

Chapter 3 ended with the two halves of every N-body code: a function that computes accelerations from positions, and a function that advances the state by one step. This chapter is about the second one. `simulate_system()` offers three choices for it, through its `method` argument: `"euler"`, `"euler_cromer"`, and `"verlet"`, the default. They differ by a few characters of code and by everything that matters. The textbook method quietly feeds energy into every orbit it touches; the default keeps a planet on a closed ellipse for as long as you care to run it. The reason is not that Verlet is "more accurate" in the usual sense. It is that Verlet respects a geometric property of Newton's equations that Euler violates, and the goal of this chapter is to make that property concrete enough that you could check for it yourself.

## The problem of taking a step

The state of a body is its position $\mathbf{x}$ and velocity $\mathbf{v}$, and the equations of motion say $\dot{\mathbf{x}} = \mathbf{v}$ and $\dot{\mathbf{v}} = \mathbf{a}(\mathbf{x})$, where the acceleration depends only on positions (Chapter 3). Given the state at time $t$, we want the state at $t + \Delta t$.

Taylor's theorem is the whole toolkit. For a smooth trajectory,

$$
\begin{aligned}
\mathbf{x}(t+\Delta t) &= \mathbf{x} + \mathbf{v}\,\Delta t + \tfrac12\,\mathbf{a}\,\Delta t^2 + \tfrac16\,\dot{\mathbf{a}}\,\Delta t^3 + O(\Delta t^4), \\
\mathbf{v}(t+\Delta t) &= \mathbf{v} + \mathbf{a}\,\Delta t + \tfrac12\,\dot{\mathbf{a}}\,\Delta t^2 + O(\Delta t^3),
\end{aligned}
$$ {#eq-taylor}

where everything on the right is evaluated at $t$. We know $\mathbf{x}$ and $\mathbf{v}$, and we can compute $\mathbf{a}$ from $\mathbf{x}$. We do not know $\dot{\mathbf{a}}$ or anything beyond it. Every integrator is a decision about what to do with the terms we cannot compute.

A method whose single-step error is $O(\Delta t^{p+1})$ is said to be of *order* $p$, because over a fixed total time $T$ there are $T/\Delta t$ steps and the errors accumulate to $O(\Delta t^p)$. Halve the step and a first-order method's error halves; a second-order method's error drops by four. That is the usual way to grade integrators, and it will turn out to be only half the story.

## Euler's method

The simplest decision is to keep only the terms we can compute exactly and drop the rest:

$$
\begin{aligned}
\mathbf{x}_{n+1} &= \mathbf{x}_n + \mathbf{v}_n\,\Delta t, \\
\mathbf{v}_{n+1} &= \mathbf{v}_n + \mathbf{a}_n\,\Delta t.
\end{aligned}
$$ {#eq-euler}

Both updates truncate @eq-taylor after the first power of $\Delta t$, so the local error is $O(\Delta t^2)$ and the method is first order. It is what every introductory text starts with, and it is `method = "euler"`.

It is also wrong for orbits in a way that halving the step does not fix. Consider a body on a circular orbit, where $\mathbf{v} \perp \mathbf{x}$ and $\mathbf{a} \perp \mathbf{v}$. After one Euler step,

$$
|\mathbf{x}_{n+1}|^2 = |\mathbf{x}_n|^2 + 2\Delta t\,\mathbf{x}_n\cdot\mathbf{v}_n + \Delta t^2 |\mathbf{v}_n|^2 = r^2 + v^2\Delta t^2,
$$

so the body ends the step slightly farther out, and

$$
|\mathbf{v}_{n+1}|^2 = v^2 + 2\Delta t\,\mathbf{v}_n\cdot\mathbf{a}_n + \Delta t^2 a^2 = v^2 + a^2\Delta t^2,
$$

so it also ends the step slightly faster. Farther out means higher potential energy; faster means higher kinetic energy. Every single step adds energy to the orbit, in both accounts at once, and there is no step that ever takes any back. The orbit spirals outward, forever. A smaller $\Delta t$ makes each step's contribution smaller, but there are correspondingly more steps, and the sign never changes.

There is a cleaner way to see what is going on, using the simplest oscillating system there is: a mass on a spring, $\ddot{x} = -\omega^2 x$, with state $(x, v)$. One Euler step is a linear map,

$$
\begin{pmatrix} x_{n+1} \\ v_{n+1} \end{pmatrix}
= \underbrace{\begin{pmatrix} 1 & \Delta t \\ -\omega^2\Delta t & 1 \end{pmatrix}}_{M_{\mathrm{E}}}
\begin{pmatrix} x_n \\ v_n \end{pmatrix},
\qquad \det M_{\mathrm{E}} = 1 + \omega^2\Delta t^2.
$$ {#eq-euler-matrix}

The determinant of a linear map is the factor by which it scales areas. The true dynamics of the oscillator carries any region of the $(x, v)$ plane around an ellipse without changing its area, a fact that holds for every Hamiltonian system and is known as Liouville's theorem. Euler's map scales area by $1 + \omega^2\Delta t^2 > 1$ on every step. The oscillator's energy is proportional to the area of the ellipse it traces in the $(x, v)$ plane, so a map that inflates areas inflates energies. Nothing about that depends on $\Delta t$ being large.

## Euler-Cromer

Here is a change so small it looks like a typo:

$$
\begin{aligned}
\mathbf{v}_{n+1} &= \mathbf{v}_n + \mathbf{a}_n\,\Delta t, \\
\mathbf{x}_{n+1} &= \mathbf{x}_n + \mathbf{v}_{n+1}\,\Delta t.
\end{aligned}
$$ {#eq-euler-cromer}

The velocity is updated first, and the position update uses the *new* velocity. This is `method = "euler_cromer"`, also called the semi-implicit Euler method or symplectic Euler. Its local error is still $O(\Delta t^2)$: it is a first-order method, no more accurate than plain Euler by the usual grading. But redo the oscillator calculation. Substituting the first line into the second, $x_{n+1} = x_n + (v_n - \omega^2 x_n \Delta t)\Delta t$, so

$$
M_{\mathrm{EC}} = \begin{pmatrix} 1 - \omega^2\Delta t^2 & \Delta t \\ -\omega^2\Delta t & 1 \end{pmatrix},
\qquad \det M_{\mathrm{EC}} = (1 - \omega^2\Delta t^2) + \omega^2\Delta t^2 = 1.
$$ {#eq-ec-matrix}

Exactly one. The Euler-Cromer map preserves area, for every $\Delta t$, just as the true dynamics does. An integrator with this property is called *symplectic*. (In one spatial dimension, symplectic means area-preserving; in more dimensions it is a slightly stronger condition that reduces to the same idea.) A symplectic integrator cannot have the systematic energy drift that Euler has, because energy drift is area growth. Its energy error oscillates, and for a small enough step it stays bounded forever.

::: {.callout-note title="Where it comes from"}
The method is named for Alan Cromer, who published it in a two-page note in the *American Journal of Physics* in 1981 [@cromer1981]. He reports that the reordering was found by one of his students, who noticed that it made the Euler method work on oscillators when it otherwise would not. The method itself had been known to numerical analysts under other names, but Cromer's note is where the physics-teaching community learned that the order of two lines of code could be the difference between a stable orbit and an unstable one.
:::

Why does this version work when the other does not? The position update in @eq-euler-cromer uses $\mathbf{v}_{n+1}$, which is the velocity at the *end* of the step. In effect it averages a forward-looking velocity with a backward-looking acceleration, and the two mistakes cancel in exactly the way needed to preserve area. There is a deeper structure here: the map is the composition of two exact flows, a "kick" that changes only the velocity and a "drift" that changes only the position, and each of those is itself symplectic because it is the exact solution of a Hamiltonian system (a free particle for the drift, a particle frozen in place for the kick). Composing symplectic maps gives a symplectic map. Plain Euler is not a composition of exact flows, and that is what it costs.

Euler-Cromer is fast, needing one acceleration evaluation per step, and is a reasonable choice for a quick preview. But first order is first order. Its positions drift in *phase* along the orbit at a rate proportional to $\Delta t$ even when the energy is right, so a planet simulated with Euler-Cromer is on the correct ellipse at the wrong time. Something better is needed for real work.

## Velocity Verlet

Go back to @eq-taylor and keep one more term. The position expansion through $\Delta t^2$ uses only quantities we have,

$$
\mathbf{x}_{n+1} = \mathbf{x}_n + \mathbf{v}_n\Delta t + \tfrac12\mathbf{a}_n\Delta t^2,
$$

with local error $O(\Delta t^3)$. For the velocity, the troublesome term is $\tfrac12\dot{\mathbf{a}}\Delta t^2$. We cannot compute $\dot{\mathbf{a}}$ directly, but having just computed $\mathbf{x}_{n+1}$, we can compute $\mathbf{a}_{n+1} = \mathbf{a}(\mathbf{x}_{n+1})$, and then approximate the derivative by the difference quotient $\dot{\mathbf{a}} \approx (\mathbf{a}_{n+1} - \mathbf{a}_n)/\Delta t$. Substituting,

$$
\mathbf{v}_{n+1} = \mathbf{v}_n + \mathbf{a}_n\Delta t + \tfrac12(\mathbf{a}_{n+1} - \mathbf{a}_n)\Delta t
= \mathbf{v}_n + \tfrac12(\mathbf{a}_n + \mathbf{a}_{n+1})\Delta t.
$$

Together:

$$
\begin{aligned}
\mathbf{x}_{n+1} &= \mathbf{x}_n + \mathbf{v}_n\,\Delta t + \tfrac12\,\mathbf{a}_n\,\Delta t^2, \\
\mathbf{a}_{n+1} &= \mathbf{a}(\mathbf{x}_{n+1}), \\
\mathbf{v}_{n+1} &= \mathbf{v}_n + \tfrac12\left(\mathbf{a}_n + \mathbf{a}_{n+1}\right)\Delta t.
\end{aligned}
$$ {#eq-verlet}

This is **Velocity Verlet**, `method = "verlet"`. Position first, then the new acceleration, then the velocity using the average of the old and new accelerations, which is exactly the recipe in the package documentation. It is second order: halving the step divides the error by four.

Second order is nice, but it is not why the method is the default. Redo the oscillator one more time. With $\mathbf{a} = -\omega^2 x$, the position update is $x_{n+1} = (1 - \tfrac12\omega^2\Delta t^2)\,x_n + \Delta t\, v_n$, and substituting into the velocity update gives, after a few lines of algebra,

$$
M_{\mathrm{VV}} = \begin{pmatrix}
1 - \tfrac12\omega^2\Delta t^2 & \Delta t \\[2pt]
-\omega^2\Delta t\left(1 - \tfrac14\omega^2\Delta t^2\right) & 1 - \tfrac12\omega^2\Delta t^2
\end{pmatrix},
\qquad \det M_{\mathrm{VV}} = 1.
$$ {#eq-vv-matrix}

(Multiply out the determinant: the $\omega^2\Delta t^2$ and $\omega^4\Delta t^4$ terms cancel exactly.) Velocity Verlet is symplectic. It is also *time-reversible*: if you take a step forward and then a step backward with the velocity reversed, you land exactly where you started, which is a symmetry the true equations of motion have and Euler and Euler-Cromer do not. Symplectic, time-reversible, and second order, at the cost of a second acceleration evaluation per step. (orbitr computes $\mathbf{a}_n$ at the top of each step and $\mathbf{a}_{n+1}$ after the position update, so a Verlet step costs two force evaluations to Euler-Cromer's one. Since $\mathbf{a}_{n+1}$ is the next step's $\mathbf{a}_n$, an implementation that kept it would pay for only one; Chapter 15 comes back to this.) For gravitational dynamics, that combination is very hard to beat.

### Three names for one method

Velocity Verlet has two other common forms, and recognizing them will save you confusion when reading other codes.

**The Størmer, or position, form.** Write the Taylor expansion of $\mathbf{x}$ both forward and backward in time and add them. The odd powers of $\Delta t$ cancel:

$$
\mathbf{x}(t+\Delta t) + \mathbf{x}(t-\Delta t) = 2\mathbf{x}(t) + \mathbf{a}(t)\Delta t^2 + O(\Delta t^4),
$$

so

$$
\mathbf{x}_{n+1} = 2\mathbf{x}_n - \mathbf{x}_{n-1} + \mathbf{a}_n\Delta t^2.
$$ {#eq-stormer}

No velocities appear at all, and the local error in position is $O(\Delta t^4)$, one order better than the derivation of @eq-verlet suggested. Subtracting consecutive position updates of @eq-verlet and using the velocity update to eliminate $\mathbf{v}$ (a few lines of algebra worth doing once) shows that velocity Verlet produces exactly the same sequence of positions as @eq-stormer. Velocity Verlet is just the Störmer form with the velocity carried along explicitly, which is convenient because you usually want the velocity.

**The leapfrog, or kick-drift-kick, form.** Split the velocity update of @eq-verlet in half, doing the first half before the position update and the second half after:

$$
\mathbf{v}_{n+\frac12} = \mathbf{v}_n + \tfrac12\mathbf{a}_n\Delta t, \qquad
\mathbf{x}_{n+1} = \mathbf{x}_n + \mathbf{v}_{n+\frac12}\Delta t, \qquad
\mathbf{v}_{n+1} = \mathbf{v}_{n+\frac12} + \tfrac12\mathbf{a}_{n+1}\Delta t.
$$ {#eq-leapfrog}

Substitute the first line into the second and you recover the position update of @eq-verlet; substitute it into the third and you recover the velocity update. The three forms are algebraically identical. The leapfrog form makes the kick-drift-kick structure explicit, and like Euler-Cromer it is a composition of exact flows, which is the structural reason it is symplectic. Astrophysicists usually say "leapfrog"; chemists say "velocity Verlet"; numerical analysts say "Störmer-Verlet." They are all talking about @eq-verlet.

::: {.callout-note title="Where it comes from"}
The method has been discovered at least four times, by people working on completely different problems.

The first was Newton. Proposition 1 of Book I of the *Principia* [@newton1687] proves Kepler's area law by imagining the central force acting as a sequence of discrete impulses: the body drifts in a straight line, receives a kick toward the center, drifts again. Newton then lets the interval between kicks go to zero. But stop before the limit and what you have is the leapfrog scheme, exactly. Hairer, Lubich, and Wanner make this point in their survey of the method [@hairer2003], and it is a satisfying one: the very first proof about orbits under gravity was, in effect, a symplectic integrator.

The second was Carl Størmer, a Norwegian mathematician who in 1907 wanted to understand the aurora [@stormer1907]. He computed the paths of charged particles in Earth's magnetic field, by hand, with a team of assistants, using the position form @eq-stormer. The name "Størmer's method" (usually spelled Störmer in the numerical-analysis literature) stuck.

The third was Loup Verlet, in 1967 [@verlet1967], simulating liquid argon as a few hundred Lennard-Jones atoms, one of the first molecular dynamics simulations. He needed a method that would keep the total energy of the fluid constant over long runs, found the position form, and the chemistry community has called it the Verlet algorithm ever since. The velocity form @eq-verlet was written down by Swope, Andersen, Berens, and Wilson in 1982 [@swope1982], in a paper about water clusters.

Somewhere in between, Richard Feynman taught the method to Caltech freshmen. Chapter 9 of the first volume of the *Lectures on Physics* [@feynman1963] integrates a planetary orbit by hand, step by step, with the velocities evaluated at the half-steps, which is the leapfrog form. He does not give it a name.

The reason the same method keeps turning up is the reason it is orbitr's default: anyone who needs long-time stability in a conservative system, and tries the obvious thing first, ends up here.
:::

### What "conserves energy" really means

Velocity Verlet does *not* conserve energy exactly. No method with a fixed step can, for a nonlinear force. What it does is better than it sounds. A symplectic integrator applied to a Hamiltonian $H$ turns out to be the *exact* solution of a slightly different Hamiltonian, $\tilde H = H + \Delta t^2 H_2 + \Delta t^4 H_4 + \cdots$, called the modified or shadow Hamiltonian. The integrator conserves $\tilde H$ exactly, step after step (more precisely, over exponentially long times), so the true energy $H$ can only wander as far from its starting value as $\tilde H$ differs from $H$, which is $O(\Delta t^2)$. The energy error oscillates within a band and never leaves it. This is called backward error analysis, and Hairer, Lubich, and Wanner's book is the standard reference [@hairer2006].

Euler has no shadow Hamiltonian, because it is not symplectic, and so its energy error is free to grow without bound. The difference between a bounded oscillation and a secular drift is the difference between a planet that is still there after a million years and one that is not.

## Seeing it in orbitr

Here is the vignette's comparison. A star and a planet, launched at 30 km/s from $10^{11}$ m, which is an ellipse with $e \approx 0.35$ and a period of about 535 days. To make the differences obvious I use a time step of six hours rather than the vignette's one hour; with hourly steps the same things happen, more slowly.

```{r}
star_planet <- create_system() |>
  add_body("Star",   mass = 1e30) |>
  add_body("Planet", mass = 1e24, x = 1e11, vy = 30000)

run <- function(method, time_step = seconds_per_hour * 6,
                duration = seconds_per_year * 10) {
  star_planet |>
    simulate_system(time_step = time_step, duration = duration,
                    method = method) |>
    mutate(method = method)
}

runs <- bind_rows(run("verlet"), run("euler_cromer"), run("euler"))
```

```{r}
#| label: fig-three-methods
#| fig-cap: "Ten years of the same orbit with three integrators and a six-hour step. Verlet retraces one ellipse; Euler-Cromer stays on an ellipse of the right size but slips in phase; Euler spirals outward."
#| fig-width: 5.5
#| fig-height: 4.6
#| out-width: "85%"
runs |>
  filter(id == "Planet") |>
  ggplot(aes(x, y, color = method)) +
  geom_path(alpha = 0.7) +
  annotate("point", x = 0, y = 0, size = 3) +
  coord_equal() +
  labs(x = NULL, y = NULL, color = NULL)
```

The picture is the one in the documentation. Now the number behind it. The total energy of any N-body system is the kinetic energy of every body plus the potential energy of every *pair*,

$$
E = \sum_j \tfrac12 m_j v_j^2 \;-\; \sum_{j<k} \frac{G m_j m_k}{r_{jk}},
$$ {#eq-total-energy}

and both are easy to compute from the output tibble. The potential energy needs every pair at every time step, which is a self-join on `time`:

```{r}
total_energy <- function(sim, G = gravitational_constant) {
  kinetic <- sim |>
    group_by(time) |>
    summarise(KE = sum(0.5 * mass * (vx^2 + vy^2 + vz^2)))

  potential <- sim |>
    select(time, id, mass, x, y, z) |>
    inner_join(sim |> select(time, id2 = id, mass2 = mass,
                             x2 = x, y2 = y, z2 = z),
               by = "time", relationship = "many-to-many") |>
    filter(id < id2) |>                       # each unordered pair once
    mutate(r = sqrt((x - x2)^2 + (y - y2)^2 + (z - z2)^2)) |>
    group_by(time) |>
    summarise(PE = -sum(G * mass * mass2 / r))

  inner_join(kinetic, potential, by = "time") |>
    mutate(E = KE + PE,
           rel_error = (E - first(E)) / abs(first(E)))
}
```

This is what the package's `get_energy()` computes, one row per time step with `kinetic`, `potential`, and `energy` columns (and it reads the softening length from the run, so the potential matches the force law that was actually used). `conserved_quantities()` goes one step further and adds the relative error $(E - E_0)/|E_0|$ as `energy_error`, together with the total momentum and angular momentum, which Chapter 12 uses. Having seen the sum written out once, I will use those from here on.

```{r}
#| label: fig-energy-drift
#| fig-cap: "Relative energy error $(E - E_0)/|E_0|$ for the three runs. Note the different vertical scales: Euler's error grows without limit, while the two symplectic methods oscillate within a fixed band, with spikes at each periapsis passage where the step is effectively largest."
#| fig-width: 5.5
#| fig-height: 5
#| out-width: "90%"
energy <- runs |>
  group_by(method) |>
  group_modify(~ conserved_quantities(.x)) |>
  ungroup()

energy |>
  ggplot(aes(time / seconds_per_year, energy_error)) +
  geom_line() +
  facet_wrap(~ method, scales = "free_y", ncol = 1) +
  labs(x = "Years", y = "Relative energy error")
```

The Euler run gains tens of percent of its binding energy in ten years and would eventually escape. The two symplectic runs show the bounded oscillation that backward error analysis promises: the error rises and falls once per orbit, biggest near periapsis where the planet moves fastest and a fixed step covers the most angle, and comes back to nearly zero each time. It does not accumulate.

### Order, measured

The last claim to check is the order. Run Euler and Verlet at four step sizes and record the worst energy error over two years:

```{r}
steps <- seconds_per_day * c(1, 0.5, 0.25, 0.125)

convergence <- bind_rows(lapply(c("euler", "verlet"), function(m) {
  bind_rows(lapply(steps, function(dt) {
    star_planet |>
      simulate_system(time_step = dt, duration = seconds_per_year * 2,
                      method = m) |>
      conserved_quantities() |>
      summarise(max_rel_error = max(abs(energy_error))) |>
      mutate(method = m, step_days = dt / seconds_per_day)
  }))
}))

convergence |>
  group_by(method) |>
  mutate(ratio = lag(max_rel_error) / max_rel_error) |>
  ungroup() |>
  select(method, step_days, max_rel_error, ratio)
```

Each halving of the step cuts Euler's error by about two and Verlet's by about four: first order and second order, as derived. The ratios are not exactly 2 and 4 because the error has higher-order pieces too, and because the "worst" error for Verlet is set by the periapsis passages, which a coarser step samples differently each time. They are close.

## Choosing a time step

Everything above gives you the tools to choose `time_step` sensibly rather than by habit.

**The fastest orbit sets the step.** The error in a step depends on how much the acceleration changes during it, which is to say on how large a fraction of an orbit the step covers. In a system with the Sun, Earth, and the Moon, the Moon's 27-day orbit, not Earth's year, decides the step. A good rule of thumb is a few hundred to a thousand steps per orbit of the tightest orbit in the system: an hour for the Moon, a day for the planets, as the package defaults suggest.

**Periapsis is where it hurts.** For an eccentric orbit the body moves fastest at periapsis, and the angular rate there is $h/r_p^2$, larger than the mean rate by a factor of roughly $(1+e)/(1-e)$ when $e$ is substantial. A comet with $e = 0.97$ moves through periapsis about sixty times faster than its average, and a step that is fine at aphelion is useless at perihelion. Chapter 13 deals with this.

**Stability is not accuracy.** For the oscillator, the Verlet map is stable (its eigenvalues stay on the unit circle) as long as $\omega\Delta t < 2$, which is only about three steps per orbit. Beyond that the solution blows up; short of it, the solution is bounded but may be badly wrong. Do not take "it didn't blow up" as evidence that the step was small enough.

**Test by halving.** The convergence table is the test. Run the simulation, halve `time_step`, run it again, and compare the quantities you care about: final positions, periods, the energy band. If the answers move by less than you care about, the step was fine. If they move a lot, halve again. This costs a factor of two or three in compute time once and saves you from publishing a numerical artifact.

::: {.callout-tip title="Key results"}
- Euler: first order, increases phase-space area every step, energy grows without bound. Educational only.
- Euler-Cromer: first order, symplectic. Energy error bounded, phase error $O(\Delta t)$. Fast previews.
- Velocity Verlet (= Störmer = leapfrog): second order, symplectic, time-reversible. Energy error bounded and $O(\Delta t^2)$. Use it.
- Total energy of an N-body system: $E = \sum_j \tfrac12 m_j v_j^2 - \sum_{j<k} G m_j m_k / r_{jk}$.
- Choose the step from the fastest orbit in the system, check it by halving.
:::
