# Softening: What Happens When Bodies Get Close {#sec-softening}

```{r}
#| include: false
source("_common.R")
```

The inverse-square force has a hole in it. As two point masses approach, $1/r^2$ grows without limit, and at $r = 0$ it is infinite. Nothing in the real universe does this, because nothing in the real universe is a point: two stars that get close enough touch, and from then on gravity is the least of their problems. But the simulation does not know that. `simulate_system()` has an argument, `softening`, whose entire purpose is to paper over the hole. This chapter explains what the argument does to the force law, where the formula comes from, what it costs you when bodies are far apart, and when you should use it. The short version: softening is a model of bodies having size, not a fix for a time step that is too large, and it is off by default for a reason.

## The singularity, numerically

An integrator with a fixed step assumes the acceleration does not change much during a step. Near a close encounter that assumption fails completely. The characteristic time of the encounter is roughly the periapsis distance divided by the periapsis speed, and for a deep encounter that can be minutes while the step is an hour. The integrator then takes one enormous kick in a direction computed from a position that is already wrong, and the result is two bodies flying apart with far more energy than they came in with. Nothing blows up to `NaN`; it just produces garbage that looks like physics.

Here is a deliberately bad encounter: two Sun-mass stars approaching each other on a path that would bring them within about two-thirds of a solar radius, which for real stars is a collision.

```{r}
M  <- 1e30
mu <- gravitational_constant * 2 * M
u  <- 1.2e5     # initial relative speed, m/s
b  <- 3e9       # initial sideways offset, m

encounter <- function(softening = 0, time_step = 600,
                      duration = 4 * seconds_per_day) {
  create_system() |>
    add_body("A", mass = M, x = -1e10, y = -b / 2, vx =  u / 2) |>
    add_body("B", mass = M, x =  1e10, y =  b / 2, vx = -u / 2) |>
    simulate_system(time_step = time_step, duration = duration,
                    softening = softening)
}
```

Before running anything, Chapter 4 tells us what should happen. From the initial relative state we can get the eccentricity, and from the eccentricity the periapsis distance and the angle through which the stars should be deflected. For a hyperbola, each asymptote makes an angle $\arccos(-1/e)$ with the periapsis direction, so the total deflection is $\delta = 2\arcsin(1/e)$:

```{r}
start <- encounter(duration = 600)        # one step: just the initial state
e0    <- get_orbital_elements(start, "B", "A", mu = mu)$e[1]   # mu = G(M_A + M_B)
r_p   <- (b * u)^2 / mu / (1 + e0)        # p / (1 + e), with h = b u

c(e = e0, periapsis_solar_radii = r_p / 6.957e8,
  deflection_deg = 2 * asin(1 / e0) * 180 / pi)
```

The stars should swing around each other through about `r round(2 * asin(1 / e0) * 180 / pi)` degrees, nearly reversing direction, and leave with exactly the eccentricity and energy they arrived with. Now run it with a ten-minute step and no softening:

```{r}
bad <- encounter(softening = 0, time_step = 600)

c(e_start = get_orbital_elements(bad, "B", "A", mu = mu)$e[1],
  e_end   = get_orbital_elements(bad, "B", "A", mu = mu)$e[nrow(bad) / 2],
  max_energy_error = max(abs(conserved_quantities(bad)$energy_error)))
```

The eccentricity, which should be a constant of the motion, is not even close, and the energy error is enormous. The time step was far too large for the periapsis passage, and the integrator did what integrators do with too large a step: it invented energy. The paths look like a slingshot because that is what a huge spurious kick looks like.

## Plummer softening

The package's remedy replaces the distance in the force law:

$$
r \;\longrightarrow\; r_{\mathrm{soft}} = \sqrt{r^2 + \varepsilon^2},
$$

so the acceleration of body $j$ due to body $k$ becomes

$$
\mathbf{a}_{jk} = G m_k \,\frac{\mathbf{r}_k - \mathbf{r}_j}{\left(r_{jk}^2 + \varepsilon^2\right)^{3/2}}.
$$ {#eq-softened}

The difference vector in the numerator is untouched; only the $r^3$ in the denominator is softened. At $r = 0$ the acceleration is now zero, not infinite, and its maximum magnitude, which occurs at $r = \varepsilon/\sqrt2$, is about $0.38\,G m_k/\varepsilon^2$.

::: {.callout-note title="Where it comes from"}
In 1911 H. C. Plummer was fitting the observed brightness profiles of globular clusters and found that a density law of the form $\rho(r) \propto (1 + r^2/\varepsilon^2)^{-5/2}$ matched them well [@plummer1911]. The gravitational potential of that density distribution works out to

$$
\Phi(r) = -\frac{GM}{\sqrt{r^2 + \varepsilon^2}},
$$

a potential that looks exactly like a point mass at large $r$ and flattens out smoothly to a finite value at the center. Differentiating, the acceleration is $-d\Phi/dr = -GMr/(r^2+\varepsilon^2)^{3/2}$, which is @eq-softened. So softening is not an ad hoc patch: it replaces each point mass by a small Plummer sphere of scale radius $\varepsilon$, and the resulting force is still exactly the gradient of a potential, which means the softened system still conserves energy (its own softened energy) and angular momentum. When N-body simulation of star clusters began in earnest in the 1960s, Aarseth and others adopted the Plummer form because it was the simplest potential that was both finite at the origin and physically interpretable [@aarseth2003], and it has been the default ever since.
:::

@fig-softened-force shows what the softened force looks like. Inside $\varepsilon$ it falls to zero; outside a few $\varepsilon$ it is indistinguishable from the point-mass force.

```{r}
#| label: fig-softened-force
#| fig-cap: "The softened acceleration (solid) compared with the point-mass $1/r^2$ law (dashed), in units where $Gm/\\varepsilon^2 = 1$. Beyond $r \\approx 3\\varepsilon$ the two agree to a few percent."
#| fig-height: 3
tibble(x = seq(0.05, 5, by = 0.01)) |>
  mutate(point_mass = 1 / x^2,
         softened   = x / (1 + x^2)^1.5) |>
  tidyr::pivot_longer(-x, names_to = "law", values_to = "a") |>
  ggplot(aes(x, a, linetype = law)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 2)) +
  labs(x = expression(r / epsilon), y = "Acceleration", linetype = NULL)
```

## What softening costs

The softened force is weaker than the true force at *every* distance, not just inside $\varepsilon$. The ratio is

$$
\frac{a_{\mathrm{soft}}}{a} = \left(1 + \frac{\varepsilon^2}{r^2}\right)^{-3/2} \approx 1 - \frac{3}{2}\frac{\varepsilon^2}{r^2}
\qquad (r \gg \varepsilon).
$$ {#eq-softening-error}

For a circular orbit the period goes as the inverse square root of the force, so softening lengthens every period by a fraction of about $\tfrac34\,\varepsilon^2/r^2$. Put numbers in. The documentation suggests `softening = 1e4` (10 km) as reasonable for planetary systems; at Earth's distance from the Sun, $\varepsilon/r \approx 7\times10^{-8}$ and the period changes by one part in $10^{14}$. You could not detect that with a million years of simulation. At $\varepsilon = 10^9$ m the change at 1 AU is $3\times10^{-5}$, about a quarter of an hour per year, which starts to matter for precise work. At $\varepsilon = 10^{10}$ m, a bit under a tenth of an AU, Earth's year would be about three days too long.

The lesson is that softening is cheap as long as $\varepsilon$ is much smaller than every orbital separation you care about, and it is never free. With `softening = 0`, the default, there is no error of this kind at all, and for planetary systems, where nothing ever gets close to anything, the default is the right choice.

## The encounter, softened

Run the same encounter with the same ten-minute step and a softening length of $3\times10^9$ m, about four solar radii, which is a crude way of saying "these stars have size":

```{r}
soft <- encounter(softening = 3e9, time_step = 600)

c(e_start = get_orbital_elements(soft, "B", "A", mu = mu)$e[1],
  e_end   = get_orbital_elements(soft, "B", "A", mu = mu)$e[nrow(soft) / 2],
  max_energy_error = max(abs(conserved_quantities(soft)$energy_error)))
```

Two things changed. First, the energy is now conserved to a fraction of a percent, because the softened force varies slowly enough during the encounter for a ten-minute step to follow it. Second, the eccentricity after the encounter is different from the eccentricity before, *and that is correct*: the stars did not interact through a point-mass force, so they are not on the point-mass hyperbola. Softening changed the physics, which is what it is for. The energy diagnostic needs the same softening, so that it computes the potential energy of the force that was actually used. `conserved_quantities()` reads the softening length from the simulation output and replaces $r$ with $\sqrt{r^2+\varepsilon^2}$ in the potential automatically; if you compute the energy yourself, you have to do the same.

The third run is the honest one: no softening, and a step small enough for the point-mass physics.

```{r}
fine <- encounter(softening = 0, time_step = 20)

c(e_start = get_orbital_elements(fine, "B", "A", mu = mu)$e[1],
  e_end   = get_orbital_elements(fine, "B", "A", mu = mu)$e[nrow(fine) / 2],
  max_energy_error = max(abs(conserved_quantities(fine)$energy_error)))
```

Now the eccentricity survives the encounter and the energy error is small. That cost thirty times as many steps as the softened run, and it is the only one of the three that answers the question "what would two point masses do?"

```{r}
#| label: fig-encounter
#| fig-cap: "The same two stars in three runs. Left: unsoftened, ten-minute step, a spurious kick at periapsis. Center: softened to $3\\times 10^9$ m, same step, a smooth but physically different encounter. Right: unsoftened, twenty-second step, the true point-mass deflection."
#| fig-width: 6
#| fig-height: 2.6
bind_rows(bad  |> mutate(run = "softening = 0, dt = 10 min"),
          soft |> mutate(run = "softening = 3e9 m, dt = 10 min"),
          fine |> mutate(run = "softening = 0, dt = 20 s")) |>
  ggplot(aes(x, y, color = id)) +
  geom_path() +
  coord_equal() +
  facet_wrap(~ run) +
  labs(x = NULL, y = NULL, color = NULL) +
  theme(axis.text = element_blank())
```

## When to use it

**Use softening when bodies have size and you care about what happens when they meet.** Stars in a cluster, a planetesimal disk, anything with many bodies and frequent close passages. Set $\varepsilon$ to something like the physical size of the bodies, or to a fraction of the typical spacing, and accept that encounters closer than $\varepsilon$ are not being modeled correctly, because they would not be point-mass encounters in reality either. In this regime a modest softening also lets you use a larger time step than the closest passage would otherwise demand, which for a thousand-body system is the difference between a run that finishes and one that does not.

**Do not use softening to rescue a bad time step in a few-body system.** If two planets, or a comet and the Sun, pass close and the energy jumps, the cure is a smaller step (Chapter 13 shows how to afford one), not a softened force that changes the answer. The only exception is when you have decided that the close approach is a collision and you do not care about the trajectory afterward; then softening is a cheap way to keep the simulation from flinging things to infinity.

**Keep $\varepsilon$ far below the smallest orbit.** @eq-softening-error is the rule: the relative error in the force is about $\tfrac32(\varepsilon/r)^2$, so a softening of one percent of the smallest orbital radius changes periods by about one part in ten thousand. Decide what precision you need and work backward.

**Account for it in diagnostics.** The softened system conserves the softened energy. `get_energy()` and `conserved_quantities()` take the softening length from the run; if you hand-compute energies, use the softened distance, or you will diagnose a drift that is not there.

::: {.callout-tip title="Key results"}
- Softening replaces $r^3$ in the force denominator with $(r^2+\varepsilon^2)^{3/2}$. The force is finite everywhere and zero at $r = 0$.
- It is the force of a Plummer sphere of scale radius $\varepsilon$: still conservative, still central.
- Relative error in the force at distance $r \gg \varepsilon$: about $\tfrac32(\varepsilon/r)^2$. Periods lengthen by about $\tfrac34(\varepsilon/r)^2$.
- Softening models bodies having size. It does not fix a time step that is too large for the physics you asked for.
:::
