# Conservation Laws as Diagnostics {#sec-diagnostics}

```{r}
#| include: false
source("_common.R")
```

A simulation can be wrong in ways that are invisible in a plot. An orbit that is an ellipse of the right size, traced at the right speed, can still be rotating slowly in its plane for no physical reason, and nothing about the picture will tell you. What *will* tell you is a set of quantities that the real system keeps constant and the simulated one may not. Chapter 5 used total energy this way; this chapter adds total momentum and total angular momentum, derives why all three are conserved for any number of bodies, explains which of them a good integrator conserves *exactly* and which only approximately, and turns them into a single function you can run on any simulation output. It ends with a worked diagnosis of a simulation that passes the energy test and is still wrong.

## Three conserved quantities, for $N$ bodies

Chapter 4 derived conservation of energy and angular momentum for two bodies. The $N$-body versions follow from two properties of gravity: it is *pairwise antisymmetric* ($\mathbf{F}_{jk} = -\mathbf{F}_{kj}$) and *central* ($\mathbf{F}_{jk}$ lies along $\mathbf{r}_k - \mathbf{r}_j$).

**Momentum.** $\mathbf{P} = \sum_j m_j\mathbf{v}_j$, and

$$
\dot{\mathbf{P}} = \sum_j m_j\mathbf{a}_j = \sum_j\sum_{k\ne j}\mathbf{F}_{jk} = \sum_{j<k}\left(\mathbf{F}_{jk} + \mathbf{F}_{kj}\right) = \mathbf{0}.
$$

Every force appears twice with opposite signs. The center of mass moves in a straight line at constant velocity, which is Chapter 2's drifting Earth.

**Angular momentum.** $\mathbf{L} = \sum_j m_j\,\mathbf{r}_j\times\mathbf{v}_j$, and since $\mathbf{v}_j\times\mathbf{v}_j = 0$,

$$
\dot{\mathbf{L}} = \sum_j \mathbf{r}_j\times m_j\mathbf{a}_j = \sum_{j<k}\left(\mathbf{r}_j\times\mathbf{F}_{jk} + \mathbf{r}_k\times\mathbf{F}_{kj}\right) = \sum_{j<k}\left(\mathbf{r}_j - \mathbf{r}_k\right)\times\mathbf{F}_{jk} = \mathbf{0},
$$

because $\mathbf{F}_{jk}$ is parallel to $\mathbf{r}_j - \mathbf{r}_k$. Antisymmetry collapses the two terms of each pair into one; centrality kills it.

**Energy.** $E = \sum_j \tfrac12 m_j v_j^2 - \sum_{j<k} G m_j m_k/r_{jk}$ (@eq-total-energy). The kinetic term's rate of change is $\sum_j m_j\mathbf{v}_j\cdot\mathbf{a}_j = \sum_{j<k}\mathbf{F}_{jk}\cdot(\mathbf{v}_j - \mathbf{v}_k)$, again by antisymmetry. With $\mathbf{F}_{jk} = G m_j m_k(\mathbf{r}_k - \mathbf{r}_j)/r_{jk}^3$ and the identity $(\mathbf{r}_k - \mathbf{r}_j)\cdot(\mathbf{v}_k - \mathbf{v}_j) = r_{jk}\dot r_{jk}$ from Chapter 4,

$$
\mathbf{F}_{jk}\cdot(\mathbf{v}_j - \mathbf{v}_k) = -\frac{G m_j m_k\,\dot r_{jk}}{r_{jk}^2},
$$

and the potential term's rate of change is $\sum_{j<k} G m_j m_k\,\dot r_{jk}/r_{jk}^2$, which cancels it pair by pair. With softening (Chapter 6), replace $r_{jk}$ by $\sqrt{r_{jk}^2+\varepsilon^2}$ in the potential and the same cancellation goes through.

::: {.callout-note title="Where it comes from"}
Each of the three laws is the shadow of a symmetry. Emmy Noether proved in 1918 that every continuous symmetry of a physical system's equations gives a conserved quantity: invariance under translation in space gives momentum, under rotation gives angular momentum, under translation in *time* gives energy. The N-body equations have all three symmetries, so they have all three laws.

A numerical integrator does not have all three. A fixed-step scheme respects translations and rotations perfectly: shift or rotate every position and velocity, run a step, and you get the shifted or rotated result, exactly. But it does not respect translation in time, because time comes in lumps of $\Delta t$. Noether's theorem then says exactly what to expect: a well-constructed integrator conserves momentum and angular momentum to rounding error, but energy only approximately. That is what the next section finds.
:::

## Which laws an integrator keeps exactly

Look at the integrators of Chapter 5 as compositions of *kicks* (velocity updates from accelerations computed at fixed positions) and *drifts* (position updates at fixed velocities).

A kick changes $\mathbf{P}$ by $\Delta t\sum_j m_j\mathbf{a}_j = \Delta t\sum_j\mathbf{F}_j = \mathbf{0}$ and $\mathbf{L}$ by $\Delta t\sum_j\mathbf{r}_j\times m_j\mathbf{a}_j = \mathbf{0}$, by the same pairwise arguments as above, *exactly*, for any $\Delta t$. A drift changes neither: $\mathbf{P}$ does not involve positions, and $\mathbf{L}$ changes by $\Delta t\sum_j m_j\mathbf{v}_j\times\mathbf{v}_j = \mathbf{0}$. So any method that is a sequence of kicks and drifts, which is Euler-Cromer (kick, drift) and Velocity Verlet (half kick, drift, half kick), conserves momentum and angular momentum to machine precision no matter how large the step.

Plain Euler is not a kick followed by a drift; it updates positions with the *old* velocities and velocities with the *old* accelerations simultaneously. Momentum still survives, but angular momentum picks up $\Delta t^2\sum_j m_j\mathbf{v}_j\times\mathbf{a}_j$ per step, which does not vanish. So the three laws grade the three methods differently:

| | Euler | Euler-Cromer | Verlet |
|:--|:--|:--|:--|
| Momentum | exact | exact | exact |
| Angular momentum | $O(\Delta t)$ drift | exact | exact |
| Energy | $O(\Delta t)$ drift | bounded, $O(\Delta t)$ | bounded, $O(\Delta t^2)$ |

: What each integrator conserves (exact means to rounding error). {#tbl-conservation}

The table is the key to reading diagnostics. A momentum or angular momentum error that grows past rounding in a Verlet run is not a time-step problem; it is a bug, in your setup or in the code. An energy error is a time-step problem (or an integrator problem), and its *shape* says which.

## Computing all three

The package has one function per quantity, and each is a single grouped sum over the output tibble. Chapter 5 wrote out the energy; momentum and angular momentum are shorter still, and this is all `get_momentum()` and `get_angular_momentum()` do:

```{r}
#| eval: false
sim |>
  group_by(time) |>
  summarise(px = sum(mass * vx), py = sum(mass * vy), pz = sum(mass * vz))

sim |>
  group_by(time) |>
  summarise(Lx = sum(mass * (y * vz - z * vy)),
            Ly = sum(mass * (z * vx - x * vz)),
            Lz = sum(mass * (x * vy - y * vx)))
```

`conserved_quantities()` joins the three into one tidy time series and adds the relative errors. For energy that is $(E - E_0)/|E_0|$ as before. Momentum is often exactly zero at the start, so its error is measured against the total $\sum_j m_j|\mathbf{v}_j|$ at $t = 0$ instead; angular momentum against $|\mathbf{L}_0|$. On Chapter 5's star and planet with the three methods:

```{r}
star_planet <- create_system() |>
  add_body("Star",   mass = 1e30) |>
  add_body("Planet", mass = 1e24, x = 1e11, vy = 30000)

runs <- bind_rows(lapply(c("verlet", "euler_cromer", "euler"), function(m) {
  star_planet |>
    simulate_system(time_step = seconds_per_hour * 6,
                    duration = seconds_per_year * 10, method = m) |>
    conserved_quantities() |>
    mutate(method = m)
}))
```

```{r}
#| label: fig-diagnostics
#| fig-cap: "Relative errors in energy, momentum, and angular momentum for ten years of the same orbit with three integrators. Momentum is conserved to rounding by all three; angular momentum by the two symplectic methods; energy by none, but Verlet's error stays in a narrow band. Note the free vertical scales."
#| fig-width: 6
#| fig-height: 5
runs |>
  select(time, method, energy_error, momentum_error, angular_momentum_error) |>
  tidyr::pivot_longer(c(energy_error, momentum_error, angular_momentum_error),
                      names_to = "quantity", values_to = "error") |>
  mutate(quantity = sub("_error$", "", quantity)) |>
  ggplot(aes(time / seconds_per_year, error)) +
  geom_line() +
  facet_grid(quantity ~ method, scales = "free_y") +
  labs(x = "Years", y = "Relative error")
```

Every panel in the table above has its picture here. The momentum row is noise at the $10^{-16}$ level for all three methods. The angular momentum row is noise for Verlet and Euler-Cromer and a steady climb for Euler. The energy row is Chapter 5's figure again.

## Reading the plots

Four shapes cover nearly everything you will see in an energy diagnostic.

**A narrow band, oscillating with the orbit.** Healthy. The band's width is the integrator's $O(\Delta t^2)$ error and halving the step should shrink it by four. Nothing to do.

**A wide band, still oscillating.** The step is too large for the tightest or most eccentric orbit in the system. The run is probably not catastrophically wrong, but the next section shows that "bounded energy error" is a weaker guarantee than it sounds. Halve the step and compare.

**A steady drift.** Either the wrong integrator (Euler), or a step so large that the symplectic integrator's shadow-Hamiltonian guarantee has broken down, which happens when $\Delta t$ approaches the periapsis timescale of some orbit (Chapter 13). Check `method`, then halve the step.

**A sudden jump.** A close encounter. Two bodies passed each other faster than the step could follow and the integrator handed one of them a spurious kick (Chapter 6). Everything after the jump is on a different orbit from everything before. Either the encounter needs a smaller step (Chapter 13's segmented runs) or it is a collision in disguise and the simulation should not continue past it.

A fifth shape is an apparent drift that is not real: you ran with `softening` and computed the energy without it. The softened force conserves the softened energy; `conserved_quantities()` reads the softening length from the run, so this only bites if you compute the energy by hand or override the argument.

## A worked diagnosis: the orbit that looked fine

Mercury's perihelion precesses by 574 arcseconds per century, of which 531 come from the other planets and 43 from general relativity. Suppose you wanted to measure the planetary part with orbitr. The first question is what the integrator does to Mercury's perihelion when there are *no* other planets, where the correct answer is zero precession. Two bodies, ten years, with daily and five-day steps:

```{r}
mercury_run <- function(step_days) {
  create_system() |>
    add_sun() |>
    add_planet("Mercury", parent = "Sun") |>
    simulate_system(time_step = seconds_per_day * step_days,
                    duration = seconds_per_year * 10)
}

perihelion_angle <- function(sim) {
  # longitude of perihelion = node + argument of perihelion, unwrapped so it
  # can be tracked through 360 degrees
  get_orbital_elements(sim, "Mercury", "Sun") |>
    mutate(omega_deg = unwrap((lan + arg_pe) * pi / 180) * 180 / pi) |>
    select(time, omega_deg)
}

runs_m <- bind_rows(lapply(c(5, 2, 1), function(s) {
  sim <- mercury_run(s)
  perihelion_angle(sim) |>
    mutate(step_days = s,
           energy_error = conserved_quantities(sim)$energy_error)
}))
```

(The longitude of perihelion, $\Omega + \omega$, is the quantity to track rather than $\omega$ alone, because for Mercury's small inclination the node is poorly defined and the two angles can trade against each other while their sum stays put.)

```{r}
#| label: fig-spurious-precession
#| fig-cap: "Left: Mercury's perihelion direction in a two-body run, which should be a flat line, at three step sizes. Right: the energy error of the same runs, which is bounded in every case. Bounded energy did not prevent the orbit from rotating."
#| fig-width: 6
#| fig-height: 3
bind_rows(
  runs_m |> group_by(step_days) |>
    transmute(time, panel = "Perihelion direction (deg)",
              value = omega_deg - first(omega_deg)),
  runs_m |> group_by(step_days) |>
    transmute(time, panel = "Relative energy error", value = energy_error)
) |>
  ungroup() |>
  ggplot(aes(time / seconds_per_year, value, color = factor(step_days))) +
  geom_line() +
  facet_wrap(~ panel, scales = "free_y") +
  labs(x = "Years", y = NULL, color = "Step (days)")
```

The energy error is bounded at every step size, exactly as advertised. And the perihelion of the five-day run rotates through tens of degrees in a decade, for no physical reason at all. The daily run rotates too, more slowly. This is *numerical precession*: the integrator's shadow Hamiltonian is not Kepler's, and the difference, though it conserves energy, does not conserve the eccentricity vector. It is the single most common artifact in orbital simulation, and the energy test is blind to it.

Measure the rate at each step:

```{r}
rates <- runs_m |>
  group_by(step_days) |>
  summarise(deg_per_year = coef(lm(omega_deg ~ I(time / seconds_per_year)))[[2]])
rates
```

Halving the step cuts the rate by about four: the artifact is $O(\Delta t^2)$, like everything else about Verlet. That scaling is also the cure, because it lets you extrapolate. The real planetary precession is $531''$ per century, or $0.0015°$ per year. To get the artifact ten times smaller than the signal you would need a step of about

```{r}
rate_1d <- rates$deg_per_year[rates$step_days == 1]
target  <- 0.0015 / 10
sqrt(target / abs(rate_1d)) * 24          # hours
```

hours. That is a smaller step than anyone would guess from looking at the orbit, and it is what Chapter 8's last exercise was getting at when it asked what limits the precision. The general lesson is worth stating plainly: an energy diagnostic tells you the integrator is stable. It does not tell you the integrator is accurate. For accuracy, find a quantity the true dynamics conserves that the integrator does not, the eccentricity vector is the usual one, and watch that.

::: {.callout-tip title="Key results"}
- $N$-body gravity conserves $\mathbf{P}$, $\mathbf{L}$, and $E$ because it is pairwise antisymmetric, central, and derived from a potential.
- Kick-drift integrators (Euler-Cromer, Verlet) conserve $\mathbf{P}$ and $\mathbf{L}$ to rounding error at any step size; a drift in either means a bug. Energy is conserved only approximately, and the shape of its error says what is wrong.
- Bounded energy error does not imply an accurate orbit. Numerical precession of the periapsis is $O(\Delta t^2)$ and invisible to the energy test; track the eccentricity vector.
:::

## Exercises {.unnumbered}

::: {#exr-ch12-euler-L}
Show that one Euler step changes the total angular momentum by $\Delta t^2\sum_j m_j\mathbf{v}_j\times\mathbf{a}_j$, and that for a circular two-body orbit this quantity is nonzero and has a constant sign. Confirm with the Euler run that $|\mathbf{L}|$ grows monotonically.
:::

::: {#exr-ch12-softening}
Run Chapter 6's softened encounter and compute `conserved_quantities()` with its default and with `softening = 0`. Describe the spurious "drift" the mismatched version shows.
:::

::: {#exr-ch12-precession-e}
Repeat the numerical-precession measurement for a planet on a circular orbit ($e = 0$) and on a very eccentric one ($e = 0.8$) with the same semi-major axis and step. How does the artifact depend on eccentricity, and why?
:::

::: {#exr-ch12-euler-cromer-precession}
Repeat it with Euler-Cromer at a one-day step. Is the numerical precession larger or smaller than Verlet's at the same step, and how does it scale when the step is halved?
:::

::: {#exr-ch12-full}
Write a function `check_run(sim)` that prints one line per conservation law saying "ok" or "suspect", using thresholds you choose from the figures in this chapter, and run it on every simulation in Chapter 8.
:::
