# Comets and Highly Eccentric Orbits {#sec-comets}

```{r}
#| include: false
source("_common.R")
```

Everything in the solar system chapter was nearly circular, and nearly circular orbits are easy: the speed hardly changes, so one time step is as good as another. A comet is the opposite case. Halley's Comet spends seventy years crawling through the outer solar system and a few months sprinting around the Sun at sixty times the speed, and a step that is wasteful at aphelion is hopeless at perihelion. This chapter sets Halley up from its orbital elements, shows exactly how and where a fixed step fails on it, and then builds the tool the package does not have: a way to take small steps only when they are needed. It closes with two orbits that are not ellipses at all.

## Halley

Halley's Comet has semi-major axis 17.8 AU, eccentricity 0.967, and an inclination of 162°, which is to say it orbits the Sun *backward*, tilted 18° from the ecliptic. Its perihelion is 0.586 AU, inside Venus's orbit; its aphelion is 35 AU, beyond Neptune's. Chapter 7's conversion handles all of that:

```{r}
AU     <- distance_earth_sun
mu_sun <- gravitational_constant * mass_sun

halley <- list(a = 17.834 * AU, e = 0.96714, i = 162.26,
               lan = 58.42, arg_pe = 111.33)

halley_system <- create_system() |>
  add_sun() |>
  add_body_keplerian("Halley", mass = 2.2e14, parent = "Sun",
                     a = halley$a, e = halley$e, i = halley$i,
                     lan = halley$lan, arg_pe = halley$arg_pe, nu = 180)
```

`nu = 180` starts the comet at aphelion, where it is moving slowest. Vis-viva gives the speeds at both ends of the orbit and Kepler's third law the period:

```{r}
q <- halley$a * (1 - halley$e)      # perihelion distance
Q <- halley$a * (1 + halley$e)      # aphelion distance

c(v_perihelion_km_s = sqrt(mu_sun * (2 / q - 1 / halley$a)) / 1e3,
  v_aphelion_km_s   = sqrt(mu_sun * (2 / Q - 1 / halley$a)) / 1e3,
  period_years      = 2 * pi * sqrt(halley$a^3 / mu_sun) / seconds_per_year)
```

Fifty-five kilometers per second at perihelion, under one at aphelion: the ratio $(1+e)/(1-e)$ from Chapter 7. The timescale that matters for the integrator is the perihelion distance divided by the perihelion speed,

```{r}
tau_peri <- q / sqrt(mu_sun * (2 / q - 1 / halley$a))
tau_peri / seconds_per_day
```

about 19 days. By Chapter 5's rule a step should be a small fraction of that, so a day is marginal and a week is wrong. Meanwhile the orbit takes 75 years, and a one-day step over 75 years is 27,000 steps for a passage that occupies a few hundred of them.

## Where a fixed step fails

Run one full orbit at four step sizes and watch the energy:

```{r}
T_halley <- 2 * pi * sqrt(halley$a^3 / mu_sun)

halley_runs <- bind_rows(lapply(c(1, 2, 4, 8), function(step_days) {
  halley_system |>
    simulate_system(time_step = seconds_per_day * step_days,
                    duration = T_halley) |>
    conserved_quantities() |>
    mutate(step_days = step_days)
}))
```

```{r}
#| label: fig-halley-energy
#| fig-cap: "Relative energy error over one 75-year orbit of Halley's Comet, started at aphelion, at four step sizes. Nothing happens for 37 years; then the comet passes perihelion and the integrator with the 8-day step hands it a large spurious kick."
#| fig-height: 3.2
halley_runs |>
  ggplot(aes(time / seconds_per_year, energy_error)) +
  geom_line() +
  facet_wrap(~ step_days, scales = "free_y",
             labeller = labeller(step_days = function(s) paste(s, "day step"))) +
  labs(x = "Years", y = "Relative energy error")
```

The shape is instructive. For the first 37 years the error is essentially zero at every step size, because far from the Sun the acceleration changes slowly and even an 8-day step is tiny compared with the local timescale of years. Then, in the few weeks around perihelion, each run makes all of its error at once. For the 1-day step the kick is small and the comet comes out on nearly the right orbit. For the 8-day step the kick is a substantial fraction of the comet's binding energy, which means the comet leaves perihelion on a *different orbit*, with a different period, and every subsequent apparition is at the wrong time.

The error is $O(\Delta t^2)$ in the usual way, as the ratios between panels show, but that is cold comfort when the only way to shrink it is to spend 27,000 steps on an orbit that needs a thousand. Two of the three possible responses are bad.

**Softening is not the answer.** It would cap the force near the Sun, which is exactly the force that is correct. Halley does not hit the Sun. Softening would change the orbit to a wrong one at every perihelion, smoothly and forever.

**A uniform small step works but wastes effort.** A 6-hour step is safely inside the perihelion timescale and costs 110,000 steps per orbit. For one comet that is fine, a minute or two of computing. For a thousand comets, or a comet with a period of a thousand years, it is not.

## Segmented runs

The right response is to take small steps near perihelion and large steps elsewhere. `simulate_system()` integrates with one fixed step, but nothing stops you from running it several times, because the last row of a simulation tibble is a complete state. `continue_simulation()` does the bookkeeping: it rebuilds the system from the final time step (`system_from_simulation()` does that part on its own, if you want to edit the system between segments), runs it onward with whatever step you give it, and appends the new rows with `time` continuing from where the old run ended. The gravitational constant, integrator, and softening carry over from the previous segment unless you say otherwise.

Each segment is an ordinary Verlet run with all of Chapter 5's guarantees; only at the switches is there any disturbance, and it is a single step's worth.

When to switch? Say we want the small step inside 5 AU. Kepler's equation, run in reverse, gives the time from perihelion to any true anomaly, and the true anomaly at $r = 5$ AU follows from the orbit equation, $\cos\nu = (p/r - 1)/e$ (the inverse-direction helper `time_since_periapsis()` is in `R/helpers.R`):

```{r}
p        <- halley$a * (1 - halley$e^2)
nu_5AU   <- acos((p / (5 * AU) - 1) / halley$e) * 180 / pi
t_inside <- time_since_periapsis(nu_5AU, halley$a, halley$e, mu_sun)

c(nu_at_5AU_deg = nu_5AU, years_inside_5AU_each_way = t_inside / seconds_per_year)
```

Halley spends about a year on each side of perihelion inside 5 AU, and 73 years outside. Three segments, then: coarse from aphelion to the 5 AU crossing, fine through perihelion, coarse back out:

```{r}
coarse <- seconds_per_day * 8
fine   <- seconds_per_hour * 6
t_out  <- T_halley / 2 - t_inside           # aphelion to the inbound crossing

segmented <- halley_system |>
  simulate_system(time_step = coarse, duration = t_out) |>
  continue_simulation(time_step = fine,   duration = 2 * t_inside) |>
  continue_simulation(time_step = coarse, duration = t_out)

n_distinct(segmented$time)
```

A few thousand steps instead of 110,000, and the energy:

```{r}
#| label: fig-segmented
#| fig-cap: "Energy error for the segmented run (8-day steps outside 5 AU, 6-hour steps inside) compared with a uniform 8-day run. The segmented run resolves the perihelion passage at a small fraction of the cost of a uniform fine step."
#| fig-height: 3
bind_rows(
  conserved_quantities(segmented) |> mutate(run = "Segmented: 8 d / 6 h / 8 d"),
  halley_runs |> filter(step_days == 8) |> mutate(run = "Uniform 8 d")
) |>
  ggplot(aes(time / seconds_per_year, energy_error, color = run)) +
  geom_line() +
  labs(x = "Years", y = "Relative energy error", color = NULL)
```

```{r}
c(segmented = max(abs(conserved_quantities(segmented)$energy_error)),
  uniform_8_day = max(abs(filter(halley_runs, step_days == 8)$energy_error)),
  uniform_1_day = max(abs(filter(halley_runs, step_days == 1)$energy_error)))
```

The segmented run beats the uniform daily run on accuracy at a fraction of its cost. The idea generalizes: for a comet with many perihelion passages, loop over them; for a system with one body that needs fine steps and many that do not, you still pay for the fine steps on every body (the package integrates them all together), which is the limitation that true adaptive-step codes remove. Chapter 16 says why they are hard to do without losing Verlet's symplectic guarantee.

::: {.callout-note title="Where it comes from"}
Halley's Comet is the reason we know comets are periodic at all. Edmond Halley, using Newton's new method, computed orbits for two dozen historical comets and noticed that the ones of 1531, 1607, and 1682 had nearly the same elements. He predicted a return in 1758, sixteen years after his own death, and when it came the question of whether comets obeyed gravity was settled. Alexis Clairaut refined the prediction to within a month by computing the perturbations of Jupiter and Saturn, by hand, over the preceding orbit, which was the first serious numerical integration in celestial mechanics and took him and two assistants most of a year.
:::

## Hyperbolic visitors

In 2017 an object was found on an orbit with $e = 1.2$. It was not bound to the Sun at all; it had come from interstellar space and was on its way back out. 'Oumuamua passed perihelion at 0.255 AU and was gone within months.

The same six elements describe a hyperbola, with one twist: its semi-major axis is *negative*. The periapsis distance $a(1-e)$ stays positive, since both factors flip sign, and with $a < 0$ and $e > 1$ the semi-latus rectum $p = a(1-e^2)$ is positive, $h = \sqrt{\mu p}$ is real, and @eq-pf-position and @eq-pf-velocity work unchanged. `add_body_keplerian()` accepts $e > 1$ when `a` is negative, provided the starting true anomaly is within the asymptotes, $|\nu| < \arccos(-1/e)$ (it refuses $e = 1$ exactly, where $a$ would be infinite):

```{r}
e_ou <- 1.1995
q_ou <- 0.2553 * AU
a_ou <- -q_ou / (e_ou - 1)                   # negative for a hyperbola

c(asymptote_deg = acos(-1 / e_ou) * 180 / pi,
  deflection_deg = 2 * asin(1 / e_ou) * 180 / pi)

oumuamua <- create_system() |>
  add_sun() |>
  add_body_keplerian("Oumuamua", mass = 1e10, parent = "Sun",
                     a = a_ou, e = e_ou, i = 122.7, lan = 24.6,
                     arg_pe = 241.8, nu = -140) |>
  simulate_system(time_step = seconds_per_hour, duration = seconds_per_year * 2.4)
```

```{r}
rel <- relative_state(oumuamua, "Oumuamua", "Sun") |>
  mutate(r = sqrt(rx^2 + ry^2 + rz^2), v = sqrt(ux^2 + uy^2 + uz^2),
         energy = v^2 / 2 - mu / r)

c(r_start_AU = first(rel$r) / AU, r_end_AU = last(rel$r) / AU,
  v_max_km_s = max(rel$v) / 1e3,
  v_perihelion_predicted = sqrt(mu_sun * (1 + e_ou) / q_ou) / 1e3,
  energy_min = min(rel$energy), energy_max = max(rel$energy))
```

The specific energy is positive and constant, the maximum speed is the vis-viva prediction at perihelion ($v_p^2 = \mu(1+e)/q$, from $a \to$ negative in @eq-vis-viva), and the eccentricity at the end is the eccentricity at the start:

```{r}
range(get_orbital_elements(oumuamua, "Oumuamua", "Sun")$e)
```

The object came in at 26 km/s relative to the Sun and left at 26 km/s, deflected through the angle $2\arcsin(1/e)$ of about 113 degrees. (The real 'Oumuamua showed a small extra acceleration away from the Sun that gravity alone cannot produce, most likely outgassing, which is a non-gravitational force of the kind Chapter 16 discusses.)

## Sungrazers

The Kreutz family of comets, hundreds of which have been seen by solar observatories, have perihelia around 0.005 AU, about one solar radius. Vis-viva at that distance gives

```{r}
R_sun <- 6.957e8
sqrt(2 * mu_sun / R_sun) / 1e3
```

kilometers per second, over six hundred, which is the escape speed from the Sun's surface, since these comets arrive on nearly parabolic orbits. Most of them do not survive. The simulation would carry them through without comment: the Sun is a point, the comet is a point, and two points can pass within a solar radius, or within a meter, and the integrator will either resolve it (with a step of minutes) or produce the spurious kick of Chapter 6. The package roadmap's `radius` argument and collision handling are the fix; until then, a Kreutz comet is a case where the physics the package implements and the physics that happens part company, and the simulation's answer should be read as "it would have hit."

::: {.callout-tip title="Key results"}
- A fixed step must resolve the perihelion timescale $q/v_p$, where $v_p^2 = \mu(1+e)/q$; for Halley that is weeks, for a sungrazer minutes.
- A too-large step makes its whole error in one perihelion kick and leaves the comet on a different orbit.
- `continue_simulation()` lets you change the step between segments; Kepler's equation tells you when to switch. Softening is never the fix.
- Hyperbolic orbits ($e > 1$, $a < 0$) work in `add_body_keplerian()`; the visitor leaves with the speed it came with, deflected by $2\arcsin(1/e)$.
:::

## Exercises {.unnumbered}

::: {#exr-ch13-two-orbits}
Extend the segmented Halley run to two full orbits by adding three more segments, and compare the time of the second perihelion with Kepler's third law.
:::

::: {#exr-ch13-switch-radius}
Repeat the segmented run switching at 2 AU and at 10 AU instead of 5 AU. Tabulate the number of steps and the maximum energy error for each. Where is the sweet spot?
:::

::: {#exr-ch13-loop}
Write a function `run_segmented(system, body, center, r_switch, step_far, step_near, duration)` that runs in chunks of one year with the coarse step (using `continue_simulation()`), checks after each chunk whether the body is inside `r_switch`, and switches steps accordingly, without using Kepler's equation at all. Test it on Halley.
:::

::: {#exr-ch13-borisov}
Comet 2I/Borisov, the second interstellar visitor, had $e = 3.36$ and $q = 2.01$ AU. Set it up, run it through perihelion, and compute its deflection angle and its speed at infinity.
:::

::: {#exr-ch13-sungrazer}
Build a Kreutz-like comet with $q = 0.005$ AU and $e = 0.9999$, start it at 2 AU inbound, and find the time step at which Verlet resolves the perihelion passage (energy error below $10^{-4}$). How many steps per day is that?
:::
