# R Basics {#sec-r-basics}

```{r}
#| include: false
source("_common.R")
```

This appendix is for the reader who has never opened R. It covers the minimum needed to run and modify the code in this book: installing the software, running code, installing packages, and the small vocabulary of functions the chapters use over and over. It is deliberately brief. R is a large language and this book uses a narrow slice of it; the last section says where to go when you want the rest, and every book listed there is free online.

If you already use R, skip this. If you have used Python or MATLAB, skim it; most of what differs is in the sections on the pipe and on tibbles.

## Installing R and RStudio

Two installs, in this order.

**R** is the language. Download it from <https://cran.r-project.org>: pick your operating system, then the "base" distribution on Windows or the `.pkg` installer on a Mac, and run the installer with its defaults. On Linux, use your distribution's package manager (`sudo apt install r-base` on Ubuntu).

**RStudio** is the editor most people use, and the one this book assumes. Download the free desktop version from <https://posit.co/download/rstudio-desktop/>. (Positron, from the same company, is a newer alternative; either works.) When you open RStudio you will see four panes. The two that matter now are the **Console** at the bottom left, where you type a line of R and get an answer, and the **Source** pane above it, where you edit files.

Try the console. Type `2 + 2` next to the `>` prompt and press Enter:

```{r}
2 + 2
```

The `[1]` means "this is the first element of the result," which matters once results have more than one element, and they usually do.

## Running code

Three ways, from quickest to most reusable.

- **Console.** Type and press Enter. Good for checking something.
- **Script.** File → New File → R Script gives you a text file ending in `.R`. Type lines of code; press Ctrl+Enter (Cmd+Enter on a Mac) to run the line the cursor is on, or select several lines and press it to run them together. Save the file and you can rerun everything tomorrow.
- **Quarto or R Markdown document.** A document that mixes text and code "chunks," which is what this book is written in. File → New File → Quarto Document. Each chunk has a small green arrow to run it. This is the right format for a lab report, because the figures and numbers are regenerated from the code every time you render.

Lines starting with `#` are comments and are ignored. Assignment uses `<-`: `x <- 3` stores the value 3 under the name `x`, and typing `x` afterward prints it. (`=` also works for assignment; `<-` is the convention and the book uses it.)

```{r}
x <- 3
x * 2
```

Everything in R is case-sensitive: `mass_earth` and `Mass_Earth` are different names, and only the first one exists.

## Packages

R's base installation is small on purpose. Almost everything useful, including orbitr, comes as a **package**, which you install once and load each session.

```{r}
#| eval: false
install.packages("orbitr")      # once, downloads and installs (needs internet)
library(orbitr)                 # every session, makes the functions available
```

`install.packages()` takes the name in quotes; `library()` takes it bare (quotes work there too). If you call a function from a package you have not loaded, R says `could not find function "simulate_system"`, and the fix is `library(orbitr)`. Installing orbitr also installs dplyr and ggplot2, which the book uses for analysis and plotting; load them the same way. The preface's "What you need" section has the full list, including how to get the development version from GitHub.

Two more things about packages. `packageVersion("orbitr")` tells you which version you have; this book needs 1.0.0 or later. And `orbitr::simulate_system()`, with the double colon, calls a function from a package without loading the whole package, which is why you will see `tidyr::pivot_longer()` in a few places in the book for a package that is used only once.

## Getting help

`?simulate_system` opens the help page for a function: what it does, every argument and its default, and runnable examples at the bottom. This is the single most useful habit to form. Every function in orbitr has one, and @sec-function-reference is a condensed index of them. For a package as a whole, its website (<https://orbit-r.com> for orbitr) has longer articles than the help pages.

When something goes wrong, read the error message from the top; the first line usually says what R could not do, and the rest is where it was when it failed. The three errors new users hit most are `could not find function` (load the package), `object 'x' not found` (you have not created `x` yet, or spelled it differently), and `unexpected symbol` or `unexpected ')'` (a typo: a missing comma, an extra bracket, mismatched quotes). RStudio's editor underlines the last kind in red before you run it.

## Functions and arguments

A function call is a name followed by parentheses with the inputs, called arguments, inside:

```{r}
sqrt(16)
round(3.14159, digits = 2)
```

Arguments can be matched by position or by name. `round(3.14159, 2)` and `round(3.14159, digits = 2)` are the same call. This book names arguments almost everywhere, because `add_body("Moon", mass = mass_moon, x = distance_earth_moon, vy = speed_moon)` is readable and the positional version is not. Arguments with defaults can be left out: `add_body()` has `x`, `y`, `z`, `vx`, `vy`, and `vz` all defaulting to 0, so you only mention the ones that are not zero.

Writing your own function is the same idea in reverse:

```{r}
circular_speed <- function(mass, r, G = gravitational_constant) {
  sqrt(G * mass / r)
}

circular_speed(mass_earth, distance_earth_moon)
```

The body runs with the arguments filled in, and the last value computed is returned. The book defines a dozen small functions like this; every one of them follows this pattern.

## The pipe

Most of the book's code is written as a chain:

```{r}
#| eval: false
create_system() |>
  add_sun() |>
  add_planet("Earth", parent = "Sun") |>
  simulate_system(time_step = seconds_per_day, duration = seconds_per_year) |>
  plot_orbits()
```

The operator `|>` is the **pipe**. It takes the value on its left and passes it as the *first argument* of the function on its right, so `x |> f(y)` means `f(x, y)`. The chain above is exactly

```{r}
#| eval: false
plot_orbits(simulate_system(add_planet(add_sun(create_system()), "Earth", parent = "Sun"),
                            time_step = seconds_per_day, duration = seconds_per_year))
```

which says the same thing inside out and is why nobody writes it that way. Read a pipe top to bottom as a sequence of steps: create a system, add the Sun, add Earth, simulate, plot. Each function in orbitr takes the system or the simulation output as its first argument precisely so that it can sit in a chain like this.

You will see `%>%` in older code and in some of the books below; it is the same idea from the magrittr package, from before the pipe was built into R. `|>` needs R 4.1 or later, which is why the preface asks for it.

## Vectors, data frames, and tibbles

A **vector** is a sequence of values of one type, built with `c()`:

```{r}
speeds <- c(1000, 1022, 1044)
speeds * 2
speeds[2]
```

Arithmetic applies to every element at once, with no loop, and `[2]` picks the second element. Most of R works this way, which is why the book's code has so few loops.

A **data frame** is a table: columns are vectors, all the same length, each with a name. A **tibble** is the modern data frame, from the tidyverse, with a tidier print method and fewer surprises. Everything orbitr returns is a tibble. The output of `simulate_system()` has one row per body per time step and the columns `time`, `id`, `mass`, `x`, `y`, `z`, `vx`, `vy`, `vz`:

```{r}
sim <- create_system() |>
  add_body("Earth", mass = mass_earth) |>
  add_body("Moon", mass = mass_moon, x = distance_earth_moon, vy = speed_moon) |>
  simulate_system(time_step = seconds_per_hour, duration = seconds_per_day * 2)

sim
```

Printing a tibble shows the first ten rows and the column types (`<chr>` is text, `<dbl>` is a number). `nrow(sim)` and `names(sim)` give the row count and the column names; `sim$x` pulls one column out as a plain vector; `sim[sim$id == "Moon", ]` picks rows by a condition. The dplyr verbs in the next section do the same things more readably, and the book uses them instead.

## The six dplyr verbs the book uses

dplyr is the tidyverse package for working with tables, and the book uses six of its functions. Each takes a tibble as its first argument (so it sits in a pipe) and returns a tibble.

```{r}
library(dplyr)

sim |>
  filter(id == "Moon") |>                      # keep rows where this is true
  mutate(r_km = sqrt(x^2 + y^2 + z^2) / 1e3) |> # add a column computed from others
  select(time, r_km) |>                        # keep only these columns
  arrange(desc(r_km)) |>                       # sort rows (desc = largest first)
  slice_head(n = 3)                            # the first few rows
```

The one that takes a moment to learn is `summarise()` with `group_by()`: split the table into groups, then collapse each group to one row.

```{r}
sim |>
  group_by(id) |>
  summarise(max_speed = max(sqrt(vx^2 + vy^2 + vz^2)),
            n_steps   = n())
```

Inside any of these, column names are used bare, without quotes or `sim$`: `filter(id == "Moon")`, not `filter(sim$id == "Moon")`. Comparisons use `==` for equals, `!=` for not equals, and `&`, `|` for and, or. That is nearly all the dplyr in the book; `inner_join()`, which lines up two tables on a shared column, appears in a few chapters and is explained where it is used.

## The ggplot2 the book uses

ggplot2 builds a plot in layers. You say which columns map to which visual properties (`aes()`), add a geometry layer that draws them, and keep adding with `+`:

```{r}
#| fig-height: 2.8
library(ggplot2)

sim |>
  filter(id == "Moon") |>
  mutate(r_km = sqrt(x^2 + y^2 + z^2) / 1e3) |>
  ggplot(aes(x = time / seconds_per_hour, y = r_km)) +
  geom_line() +
  labs(x = "Hour", y = "Distance from origin (km)")
```

Note the switch from `|>` to `+` once the plot starts: ggplot2 predates the pipe and uses `+` to add layers. The geometries the book uses are `geom_line()` (a line in order of $x$), `geom_path()` (a line in the order of the rows, which is what an orbit needs), `geom_point()`, and `annotate("point", ...)` for a single marker. `aes(color = id)` colors by a column and adds a legend automatically. `facet_wrap(~ method)` splits one plot into panels, one per value of a column. `coord_equal()` makes one meter on the $x$ axis the same length as one meter on $y$, without which circles come out as ellipses, and `scale_y_log10()` gives a logarithmic axis. `labs()` sets the labels and `theme_minimal()` the look. `plot_orbits()` returns a ggplot built this way, so anything here can be added to it.

## Where to learn more

All of these are free to read online, and all are good.

- **R for Data Science**, by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund (<https://r4ds.hadley.nz>). The standard introduction to the tidyverse: dplyr, ggplot2, tibbles, and the way of working this book assumes. If you read one other book, read this one; the first few chapters are enough.
- **Hands-On Programming with R**, by Garrett Grolemund (<https://rstudio-education.github.io/hopr/>). Shorter and more about the language itself: vectors, functions, loops, the things R4DS skips. Good if you want to understand *why* R behaves as it does.
- **ggplot2: Elegant Graphics for Data Analysis**, by Hadley Wickham, Danielle Navarro, and Thomas Lin Pedersen (<https://ggplot2-book.org>). The full account of the plotting system, from the people who wrote it. Read it when `plot_orbits()` stops being enough.
- **R Graphics Cookbook**, by Winston Chang (<https://r-graphics.org>). Recipes: "I want a plot that looks like this." The fastest route to a specific figure.
- **Advanced R**, by Hadley Wickham (<https://adv-r.hadley.nz>). Not needed for this book. Read it when you start wanting to know what a function *is*.
- The **Posit cheat sheets** (<https://posit.co/resources/cheatsheets/>). One page each for dplyr, ggplot2, and RStudio; worth printing.

For the physics side, the references in the chapters and the Further Reading section at <https://orbit-r.com> do the same job.
